---
title: "Equivalence Testing Example"
output:
  word_document: default
  html_document: default
editor_options: 
  chunk_output_type: console
---


## Purpose 

This supplementary document provides example calculations following the EFSA bee guidance framework for non-inferiority testing, accompanied by R code that:

1. **Demonstrates the statistical requirements** for one-sided α = 0.20 non-inferiority testing with a 10% reduction margin under log-normal assumptions with realistic variance (CV = 20%)

2. **Computes confidence intervals** at multiple sample sizes to show when non-inferiority can or cannot be demonstrated for small observed effects (5% reduction)

3. **Illustrates the sample size asymmetry** between:
   - **Non-inferiority testing** (proving an effect is not meaningfully harmful) 
   - **Difference testing** (detecting that an effect is significantly different from control)

4. **Provides comprehensive sample size tables** comparing:
   - **5% reduction scenario**: appropriate for non-inferiority testing to demonstrate NO meaningful harm (effect is within acceptable margin)
   - **15% reduction scenario**: appropriate for difference testing to demonstrate significant harm exists
   
5. **Clarifies the distinction** between minimum sample sizes (which just meet the α criterion) and power-adjusted sample sizes (which ensure adequate statistical power for reliable conclusions)

The analysis shows that under EFSA's lenient α = 0.20 framework, demonstrating non-inferiority for a 5% reduction still requires substantially larger samples (~77 per arm for 80% power) than conventional difference testing for a 15% reduction (~24 per arm). This reflects the fundamental statistical challenge: proving safety (ruling out harm beyond a margin) is more demanding than detecting harm.

## Mapping between α and CIs

- One-sided non-inferiority at level α against a −10% margin on a ratio scale is equivalent to requiring the lower bound of the two-sided CI at level 1 − 2α to exceed 0.90 (i.e., no more than 10% reduction).
- With α = 0.2, that CI is two-sided 60%. So you're effectively checking the lower bound of a 60% CI against the 0.90 margin.


Equivalence (non-inferiority) criterion: 
Reject $H0: θ ≤ \ln(0.90)$  in favor of $H1: θ > \ln(0.90)$ 
if  $L_{60\%\,CI}(θ) > \ln(0.90)$, 
where $θ = \ln(\text{treated/control})$.


—-- 

## Worked example design and CIs under a log-normal assumption

Assumptions (parameterized in code below):
- Control mean = 20,000 organisms (e.g., colony size)
- Treated mean = 5% reduction → 19,000; ratio = 0.95
- Coefficient of variation (CV) = 20%
- Log-normal model with equal CV across arms
- Replicates per arm nC = nT (we'll explore n = 6, 12, 20)

Mapping CV to the log scale:
$$
\sigma^2 = \ln(1+\text{CV}^2), \quad \sigma = \sqrt{\ln(1+\text{CV}^2)}.
$$
On the log scale, if X_C, X_T are normal with common variance σ^2, the standard error for the difference in means is:
$$
SE(\hat{\theta}) = \sigma \sqrt{\frac{1}{n_T} + \frac{1}{n_C}},\quad \hat{\theta} = \ln(0.95).
$$

Two-sided CI for θ at level 1 − 2α:
$$
\hat{\theta} \pm z_{1-\alpha}\, SE(\hat{\theta}),
$$
transform back with exp() to obtain the CI for the ratio treated/control. Compare the CI lower bound to 0.90.

— 

## R code to compute 60% and 95% CIs, and the sample size needed

```{r}
# Inputs
mean_control <- 20000
ratio_true    <- 0.95           # 5% reduction
mean_treated  <- mean_control * ratio_true
cv            <- 0.20           # 20% CV (on original scale)
alpha_one_sided <- 0.20         # EFSA example for non-inferiority
margin_ratio    <- 0.90         # 10% reduction margin

# Helper: log-scale sigma from CV
sigma_log <- sqrt(log(1 + cv^2))

# CI function for the log ratio under equal-variance log-normal model
ci_ratio_lognormal <- function(nC, nT, ratio_hat, sigma_log, conf_level_two_sided) {
  theta_hat <- log(ratio_hat)
  alpha_two_sided <- 1 - conf_level_two_sided
  z <- qnorm(1 - alpha_two_sided/2)   # two-sided critical value
  SE <- sigma_log * sqrt(1/nC + 1/nT)
  lower <- exp(theta_hat - z * SE)
  upper <- exp(theta_hat + z * SE)
  c(lower = lower, upper = upper)
}

# Non-inferiority pass/fail at one-sided alpha (equivalently 2-sided CI with 1 - 2*alpha)
noninferiority_pass <- function(nC, nT, ratio_hat, sigma_log, margin_ratio, alpha_one_sided) {
  conf_two_sided <- 1 - 2*alpha_one_sided
  ci <- ci_ratio_lognormal(nC, nT, ratio_hat, sigma_log, conf_two_sided)
  list(pass = (ci["lower"] > margin_ratio),
       ci_lower = ci["lower"],
       ci_upper = ci["upper"],
       conf_level = conf_two_sided)
}

# Sample size per arm required to pass non-inferiority (approximate, equal n)
# Solve for n in: log(ratio_hat) - z * sigma * sqrt(2/n) > log(margin)
sample_size_noninf_equal_n <- function(ratio_hat, margin_ratio, sigma_log, alpha_one_sided) {
  z <- qnorm(1 - alpha_one_sided)     # one-sided
  gap <- log(ratio_hat) - log(margin_ratio)  # must be positive to be feasible
  if (gap <= 0) return(Inf)
  n_per_arm <- ceiling( 2 * ( (z * sigma_log) / gap )^2 )
  n_per_arm
}

# Demonstration for n = 6, 12, 20 per arm
for (n in c(6, 12, 20)) {
  ci60 <- ci_ratio_lognormal(n, n, ratio_true, sigma_log, conf_level_two_sided = 0.60)
  ci95 <- ci_ratio_lognormal(n, n, ratio_true, sigma_log, conf_level_two_sided = 0.95)
  res_noninf <- noninferiority_pass(n, n, ratio_true, sigma_log, margin_ratio, alpha_one_sided)
  cat(sprintf("\n\n== nC = nT = %d ==\n\n", n))
  cat(sprintf("60%% two-sided CI for ratio: [%.3f, %.3f]\n\n", ci60["lower"], ci60["upper"]))
  cat(sprintf("95%% two-sided CI for ratio: [%.3f, %.3f]\n\n", ci95["lower"], ci95["upper"]))
  cat(sprintf("Non-inferiority (one-sided alpha = %.1f, margin = %.2f): %s; lower=%.3f (two-sided CI level = %.0f%%)\n\n",
              alpha_one_sided, margin_ratio, ifelse(res_noninf$pass, "PASS", "FAIL"),
              res_noninf$ci_lower, 100*res_noninf$conf_level))
}

# What n per arm is needed (approximate) to PASS non-inferiority with alpha=0.2 (60% CI)?
n_needed_60 <- sample_size_noninf_equal_n(ratio_true, margin_ratio, sigma_log, alpha_one_sided)
# And with a more stringent 95% two-sided CI (equiv. alpha_one_sided = 0.025)?
n_needed_95 <- sample_size_noninf_equal_n(ratio_true, margin_ratio, sigma_log, alpha_one_sided = 0.025)
cat(sprintf("\n\nApproximate n per arm needed to pass at alpha=0.20 (60%% CI): %s\n\n", n_needed_60))
cat(sprintf("Approximate n per arm needed to pass at ~95%% two-sided criterion: %s\n\n", n_needed_95))
```

The results show: 

- With CV = 20% and an observed 5% reduction (ratio = 0.95), the 60% CI is still often too wide at modest n.
- For example, with n = 6 per arm, the 60% CI lower bound is typically below 0.90, so non-inferiority (α = 0.2; margin = 10%) fails.
- The approximate n per arm required to pass at α = 0.2 (60% CI) is about 20 (per arm) under these assumptions; at a 95% two-sided standard it exceeds 100 per arm. This illustrates that even with a lenient α, demonstrating "≤10% reduction" can be empirically challenging at realistic replication.


Using a one-sided α = 0.2 non-inferiority test with a 10% reduction margin is equivalent to requiring the lower bound of a two-sided 60% CI for the treated/control ratio to exceed 0.90. Under a log-normal model with CV = 20% and modest replication, an observed 5% reduction (ratio = 0.95) typically fails this test unless n ≈ 20 per group or larger; at conventional 95% CI levels the replication burden is substantially higher.

This example highlights that the prescribed α and margin may not be empirically verifiable under typical variance and replication, motivating feasibility assessment alongside methodological selection.

# Comparing Difference Test and Equivalence Test

## Considering alpha only vs Real Power

**Original formula (gives ~20):** Only ensures the test *can* reject H0 at α level when data are exactly as expected, but doesn't specify power:

$$n = 2 \left(\frac{z_\alpha \cdot \sigma}{\delta}\right)^2$$

**Corrected formula (gives ~77):** Includes power specification (standard for sample size calculations):

$$n = 2 \left(\frac{(z_\alpha + z_\beta) \cdot \sigma}{\delta}\right)^2$$

where:
- $z_\alpha = \text{qnorm}(1-\alpha)$ for one-sided test
- $z_\beta = \text{qnorm}(\text{power})$ 
- $\delta = \ln(0.95) - \ln(0.90) \approx 0.0541$

For α = 0.20, power = 0.80:
- $z_{0.20} \approx 0.842$
- $z_{0.80} \approx 0.842$
- Without power: $n = 2 \times (0.842 \times 0.198 / 0.0541)^2 \approx 19$
- With power: $n = 2 \times (1.684 \times 0.198 / 0.0541)^2 \approx 76$


## Summary of Design and Key Results

### Study Design
**Model**: Log-normal with equal variance; CV = 20%; equal sample sizes per arm  
**Scenarios**: 
- **5% reduction** (ratio = 0.95): goal is to demonstrate NO meaningful harm (non-inferiority)
- **15% reduction** (ratio = 0.85): goal is to detect significant harm (difference test)

**Non-inferiority margin**: 10% (ratio ≥ 0.90)  
**Test types**: One-sided non-inferiority (α = 0.05 or 0.20) vs. two-sided difference test (α = 0.05 or 0.20)

---

### Key Findings

#### 1. **5% Reduction Scenario** (proving safety)
**Appropriate test**: Non-inferiority  

| Alpha | Power | n per arm | Total n | Notes |
|-------|-------|-----------|---------|-------|
| **0.20** (EFSA) | 80% | **77** | **154** | **Recommended for EFSA compliance** |
| 0.20 (EFSA) | 90% | 121 | 242 | Higher confidence |
| 0.05 (stringent) | 80% | 166 | 332 | 2.2× more samples than EFSA |

**Minimum n (no power)**: 20 per arm (just meets α = 0.20 criterion)

#### 2. **15% Reduction Scenario** (detecting harm)
**Appropriate test**: Difference test  

| Alpha | Power | n per arm | Total n | Notes |
|-------|-------|-----------|---------|-------|
| **0.05** | **80%** | **24** | **48** | **Conventional approach** |
| 0.05 | 90% | 32 | 64 | Higher confidence |
| 0.20 | 80% | 14 | 28 | More lenient alternative |

**Minimum n (no power)**: 12 per arm (α = 0.05)

---

### Critical Comparisons

**1. Non-inferiority vs. Difference Testing**  
- Proving safety (5%, non-inferiority, α=0.20): **77 per arm**  
- Detecting harm (15%, difference test, α=0.05): **24 per arm**  
- **Non-inferiority requires 3.2× more samples** despite more lenient α

**2. Impact of Alpha Level** (for non-inferiority, 5% reduction, 80% power)  
- α = 0.20 (EFSA): **77 per arm**  
- α = 0.05 (stringent): **166 per arm**  
- **Reducing α from 0.20→0.05 increases sample size by 116%**

**3. Minimum vs. Power-adjusted Sample Sizes**  
- **Minimum n** (just meets α criterion): typically 25-35% of power-adjusted n
- **Power-adjusted n** (ensures 80% power): recommended for proper study design
- For EFSA framework: minimum n = 20, but **n = 77 needed for 80% power**



### Practical Implications

-**EFSA framework (α=0.20, one-sided, 10% margin)** is more lenient than conventional testing  
-**But still demanding**: requires ~75-80 colonies per arm for 80% power with CV=20%  
-**Proving safety is harder than detecting harm**: non-inferiority needs 3-4× more samples than difference testing  
-**Minimum sample sizes (no power)** are inadequate—power-adjusted calculations essential  
-**Higher CV or narrower margins** would increase sample size requirements further

**Overall conculusion**: Field studies should be designed with n≈75-100 per arm minimum to reliably demonstrate non-inferiority under EFSA guidance when CV=20% and true effect is ≤5%.


## R Code and Result Table

<!---```{r echo=FALSE,results='asis'} --->
```{r echo=TRUE,results='hide'}
# ============================================================================
# EFSA Bee Guidance: Non-inferiority vs. Difference Testing
# COMPLETE VERSION: Minimum vs. Power-adjusted sample sizes for both test types
# ============================================================================

library(knitr)

# ============================================================================
# INPUT PARAMETERS
# ============================================================================
mean_control <- 20000
cv            <- 0.20           # 20% CV (on original scale)
margin_ratio    <- 0.90         # 10% reduction margin

# Helper: log-scale sigma from CV
sigma_log <- sqrt(log(1 + cv^2))

cat(sprintf("\nLog-scale sigma: %.4f (CV = %.0f%%)\n\n", sigma_log, cv*100))
cat(sprintf("Margin on log scale: ln(0.90) = %.4f\n\n", log(margin_ratio)))
cat(sprintf("Effect (5%% reduction) on log scale: ln(0.95) = %.4f\n\n", log(0.95)))
cat(sprintf("Gap for non-inferiority: %.4f\n\n", log(0.95) - log(margin_ratio)))

# ============================================================================
# CI FUNCTION FOR LOG RATIO (LOG-NORMAL MODEL)
# ============================================================================
ci_ratio_lognormal <- function(nC, nT, ratio_hat, sigma_log, conf_level_two_sided) {
  theta_hat <- log(ratio_hat)
  alpha_two_sided <- 1 - conf_level_two_sided
  z <- qnorm(1 - alpha_two_sided/2)   # two-sided critical value
  SE <- sigma_log * sqrt(1/nC + 1/nT)
  lower <- exp(theta_hat - z * SE)
  upper <- exp(theta_hat + z * SE)
  c(lower = lower, upper = upper)
}

# ============================================================================
# NON-INFERIORITY TEST (ONE-SIDED α)
# ============================================================================
noninferiority_pass <- function(nC, nT, ratio_hat, sigma_log, margin_ratio, alpha_one_sided) {
  conf_two_sided <- 1 - 2*alpha_one_sided
  ci <- ci_ratio_lognormal(nC, nT, ratio_hat, sigma_log, conf_two_sided)
  list(pass = (ci["lower"] > margin_ratio),
       ci_lower = ci["lower"],
       ci_upper = ci["upper"],
       conf_level = conf_two_sided)
}

# ============================================================================
# DIFFERENCE TEST (TWO-SIDED)
# ============================================================================
difference_test <- function(nC, nT, ratio_hat, sigma_log, alpha_two_sided = 0.05) {
  theta_hat <- log(ratio_hat)
  SE <- sigma_log * sqrt(1/nC + 1/nT)
  z_stat <- theta_hat / SE
  p_value <- 2 * pnorm(-abs(z_stat))  # two-sided p-value
  reject <- p_value < alpha_two_sided
  
  ci <- ci_ratio_lognormal(nC, nT, ratio_hat, sigma_log, 1 - alpha_two_sided)
  
  list(reject = reject,
       p_value = p_value,
       z_stat = z_stat,
       ci_lower = ci["lower"],
       ci_upper = ci["upper"])
}

# ============================================================================
# SAMPLE SIZE FOR NON-INFERIORITY - VERSION 1: MINIMUM (NO POWER)
# This gives the MINIMUM n where the test CAN reject at α level if data = expectation
# Formula: n = 2 * (z_α * σ / δ)²
# ============================================================================
sample_size_noninf_minimum <- function(ratio_hat, margin_ratio, sigma_log, alpha_one_sided) {
  z_alpha <- qnorm(1 - alpha_one_sided)     # one-sided
  gap <- log(ratio_hat) - log(margin_ratio)  # must be positive to be feasible
  if (gap <= 0) return(Inf)
  n_per_arm <- ceiling( 2 * ( (z_alpha * sigma_log) / gap )^2 )
  n_per_arm
}

# ============================================================================
# SAMPLE SIZE FOR NON-INFERIORITY - VERSION 2: WITH POWER
# This is the STANDARD formula for proper study design
# Formula: n = 2 * ((z_α + z_β) * σ / δ)²
# ============================================================================
sample_size_noninf_with_power <- function(ratio_hat, margin_ratio, sigma_log, alpha_one_sided, power = 0.80) {
  z_alpha <- qnorm(1 - alpha_one_sided)     # one-sided
  z_beta <- qnorm(power)
  gap <- log(ratio_hat) - log(margin_ratio)  # must be positive to be feasible
  if (gap <= 0) return(Inf)
  n_per_arm <- ceiling( 2 * ( ((z_alpha + z_beta) * sigma_log) / gap )^2 )
  n_per_arm
}

# ============================================================================
# SAMPLE SIZE FOR DIFFERENCE TEST - VERSION 1: MINIMUM (NO POWER)
# This gives the MINIMUM n where the test CAN reject at α level if data = expectation
# Formula: n = 2 * (z_{α/2} * σ / |effect|)²
# ============================================================================
sample_size_diff_minimum <- function(ratio_hat, sigma_log, alpha_two_sided = 0.05) {
  z_alpha <- qnorm(1 - alpha_two_sided/2)   # two-sided
  effect <- abs(log(ratio_hat))  # effect size on log scale
  if (effect <= 0) return(Inf)
  n_per_arm <- ceiling( 2 * ( (z_alpha * sigma_log) / effect )^2 )
  n_per_arm
}

# ============================================================================
# SAMPLE SIZE FOR DIFFERENCE TEST - VERSION 2: WITH POWER
# This is the STANDARD formula for proper study design
# Formula: n = 2 * ((z_{α/2} + z_β) * σ / |effect|)²
# ============================================================================
sample_size_diff_with_power <- function(ratio_hat, sigma_log, alpha_two_sided = 0.05, power = 0.80) {
  z_alpha <- qnorm(1 - alpha_two_sided/2)   # two-sided
  z_beta <- qnorm(power)
  effect <- abs(log(ratio_hat))  # effect size on log scale
  if (effect <= 0) return(Inf)
  n_per_arm <- ceiling( 2 * ( ((z_alpha + z_beta) * sigma_log) / effect )^2 )
  n_per_arm
}

# ============================================================================
# DEMONSTRATION: 5% REDUCTION (ratio = 0.95)
# ============================================================================
cat("\n\n")
cat("============================================================\n\n")
cat("SCENARIO 1: 5% REDUCTION (ratio = 0.95)\n\n")
cat("Goal: Show NO meaningful difference from control\n\n")
cat("============================================================\n\n")

ratio_5pct <- 0.95

for (n in c(6, 12, 20, 30, 40, 80)) {
  ci60 <- ci_ratio_lognormal(n, n, ratio_5pct, sigma_log, conf_level_two_sided = 0.60)
  ci90 <- ci_ratio_lognormal(n, n, ratio_5pct, sigma_log, conf_level_two_sided = 0.90)
  ci95 <- ci_ratio_lognormal(n, n, ratio_5pct, sigma_log, conf_level_two_sided = 0.95)
  
  res_noninf_20 <- noninferiority_pass(n, n, ratio_5pct, sigma_log, margin_ratio, alpha_one_sided = 0.20)
  res_noninf_05 <- noninferiority_pass(n, n, ratio_5pct, sigma_log, margin_ratio, alpha_one_sided = 0.05)
  res_diff_05 <- difference_test(n, n, ratio_5pct, sigma_log, alpha_two_sided = 0.05)
  res_diff_20 <- difference_test(n, n, ratio_5pct, sigma_log, alpha_two_sided = 0.20)
  
  cat(sprintf("n = %d per arm:\n", n))
  cat(sprintf("  Non-inferiority α=0.20 (60%% CI): %s | Lower=%.3f vs margin=%.2f\n",
              ifelse(res_noninf_20$pass, "PASS  ", "FAIL  "),
              res_noninf_20$ci_lower, margin_ratio))
  cat(sprintf("  Non-inferiority α=0.05 (90%% CI): %s | Lower=%.3f vs margin=%.2f\n\n",
              ifelse(res_noninf_05$pass, "PASS  ", "FAIL  "),
              res_noninf_05$ci_lower, margin_ratio))
  cat(sprintf("  Difference test α=0.05 (95%% CI): %s | p=%.3f | CI=[%.3f, %.3f]\n\n",
              ifelse(res_diff_05$reject, "REJECT", "ACCEPT"),
              res_diff_05$p_value, res_diff_05$ci_lower, res_diff_05$ci_upper))
  cat(sprintf("  Difference test α=0.20 (80%% CI): %s | p=%.3f | CI=[%.3f, %.3f]\n\n",
              ifelse(res_diff_20$reject, "REJECT", "ACCEPT"),
              res_diff_20$p_value, res_diff_20$ci_lower, res_diff_20$ci_upper))
  cat("\n\n")
}

# ============================================================================
# DEMONSTRATION: 15% REDUCTION (ratio = 0.85)
# ============================================================================
cat("\n\n")
cat("==============================================================\n\n")
cat("SCENARIO 2: 15% REDUCTION (ratio = 0.85)\n\n")
cat("Goal: Show meaningful difference from control\n\n")
cat("==============================================================\n\n")

ratio_15pct <- 0.85

for (n in c(6, 12, 20, 30)) {
  ci60 <- ci_ratio_lognormal(n, n, ratio_15pct, sigma_log, conf_level_two_sided = 0.60)
  ci90 <- ci_ratio_lognormal(n, n, ratio_15pct, sigma_log, conf_level_two_sided = 0.90)
  ci95 <- ci_ratio_lognormal(n, n, ratio_15pct, sigma_log, conf_level_two_sided = 0.95)
  
  res_noninf_20 <- noninferiority_pass(n, n, ratio_15pct, sigma_log, margin_ratio, alpha_one_sided = 0.20)
  res_noninf_05 <- noninferiority_pass(n, n, ratio_15pct, sigma_log, margin_ratio, alpha_one_sided = 0.05)
  res_diff_05 <- difference_test(n, n, ratio_15pct, sigma_log, alpha_two_sided = 0.05)
  res_diff_20 <- difference_test(n, n, ratio_15pct, sigma_log, alpha_two_sided = 0.20)
  
  cat(sprintf("n = %d per arm:\n\n", n))
  cat(sprintf("  Non-inferiority α=0.20 (60%% CI): %s | Lower=%.3f vs margin=%.2f\n\n",
              ifelse(res_noninf_20$pass, "PASS  ", "FAIL  "),
              res_noninf_20$ci_lower, margin_ratio))
  cat(sprintf("  Non-inferiority α=0.05 (90%% CI): %s | Lower=%.3f vs margin=%.2f\n\n",
              ifelse(res_noninf_05$pass, "PASS  ", "FAIL  "),
              res_noninf_05$ci_lower, margin_ratio))
  cat(sprintf("  Difference test α=0.05 (95%% CI): %s | p=%.3f | CI=[%.3f, %.3f]\n\n",
              ifelse(res_diff_05$reject, "REJECT", "ACCEPT"),
              res_diff_05$p_value, res_diff_05$ci_lower, res_diff_05$ci_upper))
  cat(sprintf("  Difference test α=0.20 (80%% CI): %s | p=%.3f | CI=[%.3f, %.3f]\n\n",
              ifelse(res_diff_20$reject, "REJECT", "ACCEPT"),
              res_diff_20$p_value, res_diff_20$ci_lower, res_diff_20$ci_upper))
  cat("\n\n")
}

# ============================================================================
# COMPREHENSIVE SAMPLE SIZE COMPARISON TABLE
# ============================================================================
cat("\n\n")
cat("==============================================================\n\n")
cat("COMPREHENSIVE SAMPLE SIZE COMPARISON\n\n")
cat("Minimum n: Uses only z_α (just meets significance criterion)\n\n")
cat("Power-adjusted n: Uses z_α + z_β (ensures specified power)\n\n")
cat("==============================================================\n\n")

# Calculate required sample sizes for 5% reduction
# NON-INFERIORITY
n_noninf_5pct_a20_min <- sample_size_noninf_minimum(ratio_5pct, margin_ratio, sigma_log, alpha_one_sided = 0.20)
n_noninf_5pct_a05_min <- sample_size_noninf_minimum(ratio_5pct, margin_ratio, sigma_log, alpha_one_sided = 0.05)

n_noninf_5pct_a20_80 <- sample_size_noninf_with_power(ratio_5pct, margin_ratio, sigma_log, alpha_one_sided = 0.20, power = 0.80)
n_noninf_5pct_a20_90 <- sample_size_noninf_with_power(ratio_5pct, margin_ratio, sigma_log, alpha_one_sided = 0.20, power = 0.90)
n_noninf_5pct_a05_80 <- sample_size_noninf_with_power(ratio_5pct, margin_ratio, sigma_log, alpha_one_sided = 0.05, power = 0.80)
n_noninf_5pct_a05_90 <- sample_size_noninf_with_power(ratio_5pct, margin_ratio, sigma_log, alpha_one_sided = 0.05, power = 0.90)

# DIFFERENCE TEST (5%)
n_diff_5pct_a05_min <- sample_size_diff_minimum(ratio_5pct, sigma_log, alpha_two_sided = 0.05)
n_diff_5pct_a20_min <- sample_size_diff_minimum(ratio_5pct, sigma_log, alpha_two_sided = 0.20)

n_diff_5pct_a05_80 <- sample_size_diff_with_power(ratio_5pct, sigma_log, alpha_two_sided = 0.05, power = 0.80)
n_diff_5pct_a20_80 <- sample_size_diff_with_power(ratio_5pct, sigma_log, alpha_two_sided = 0.20, power = 0.80)

# Calculate required sample sizes for 15% reduction
# DIFFERENCE TEST (15%)
n_diff_15pct_a05_min <- sample_size_diff_minimum(ratio_15pct, sigma_log, alpha_two_sided = 0.05)
n_diff_15pct_a20_min <- sample_size_diff_minimum(ratio_15pct, sigma_log, alpha_two_sided = 0.20)

n_diff_15pct_a05_80 <- sample_size_diff_with_power(ratio_15pct, sigma_log, alpha_two_sided = 0.05, power = 0.80)
n_diff_15pct_a05_90 <- sample_size_diff_with_power(ratio_15pct, sigma_log, alpha_two_sided = 0.05, power = 0.90)
n_diff_15pct_a20_80 <- sample_size_diff_with_power(ratio_15pct, sigma_log, alpha_two_sided = 0.20, power = 0.80)
n_diff_15pct_a20_90 <- sample_size_diff_with_power(ratio_15pct, sigma_log, alpha_two_sided = 0.20, power = 0.90)

# Create comprehensive comparison table
comparison_table <- data.frame(
  Scenario = c(
    # 5% reduction - Non-inferiority tests (MINIMUM)
    "5% reduction",
    "5% reduction",
    # 5% reduction - Non-inferiority tests (WITH POWER)
    "5% reduction",
    "5% reduction",
    "5% reduction",
    "5% reduction",
    # 5% reduction - Difference tests (MINIMUM)
    "5% reduction",
    "5% reduction",
    # 5% reduction - Difference tests (WITH POWER)
    "5% reduction",
    "5% reduction",
    # 15% reduction - Difference tests (MINIMUM)
    "15% reduction",
    "15% reduction",
    # 15% reduction - Difference tests (WITH POWER)
    "15% reduction",
    "15% reduction",
    "15% reduction",
    "15% reduction"
  ),
  
  Objective = c(
    "Show NO meaningful harm",
    "Show NO meaningful harm",
    "Show NO meaningful harm",
    "Show NO meaningful harm",
    "Show NO meaningful harm",
    "Show NO meaningful harm",
    "Detect difference (wrong goal)",
    "Detect difference (wrong goal)",
    "Detect difference (wrong goal)",
    "Detect difference (wrong goal)",
    "Detect significant harm",
    "Detect significant harm",
    "Detect significant harm",
    "Detect significant harm",
    "Detect significant harm",
    "Detect significant harm"
  ),
  
  Test_Type = c(
    "Non-inferiority",
    "Non-inferiority",
    "Non-inferiority",
    "Non-inferiority",
    "Non-inferiority",
    "Non-inferiority",
    "Difference test",
    "Difference test",
    "Difference test",
    "Difference test",
    "Difference test",
    "Difference test",
    "Difference test",
    "Difference test",
    "Difference test",
    "Difference test"
  ),
  
  Alpha = c(
    "α=0.20 (one-sided)",
    "α=0.05 (one-sided)",
    "α=0.20 (one-sided)",
    "α=0.20 (one-sided)",
    "α=0.05 (one-sided)",
    "α=0.05 (one-sided)",
    "α=0.05 (two-sided)",
    "α=0.20 (two-sided)",
    "α=0.05 (two-sided)",
    "α=0.20 (two-sided)",
    "α=0.05 (two-sided)",
    "α=0.20 (two-sided)",
    "α=0.05 (two-sided)",
    "α=0.05 (two-sided)",
    "α=0.20 (two-sided)",
    "α=0.20 (two-sided)"
  ),
  
  CI_Level = c(
    "60%",
    "90%",
    "60%",
    "60%",
    "90%",
    "90%",
    "95%",
    "80%",
    "95%",
    "80%",
    "95%",
    "80%",
    "95%",
    "95%",
    "80%",
    "80%"
  ),
  
  Power = c(
    "—",
    "—",
    "80%",
    "90%",
    "80%",
    "90%",
    "—",
    "—",
    "80%",
    "80%",
    "—",
    "—",
    "80%",
    "90%",
    "80%",
    "90%"
  ),
  
  n_per_arm = c(
    n_noninf_5pct_a20_min,
    n_noninf_5pct_a05_min,
    n_noninf_5pct_a20_80,
    n_noninf_5pct_a20_90,
    n_noninf_5pct_a05_80,
    n_noninf_5pct_a05_90,
    n_diff_5pct_a05_min,
    n_diff_5pct_a20_min,
    n_diff_5pct_a05_80,
    n_diff_5pct_a20_80,
    n_diff_15pct_a05_min,
    n_diff_15pct_a20_min,
    n_diff_15pct_a05_80,
    n_diff_15pct_a05_90,
    n_diff_15pct_a20_80,
    n_diff_15pct_a20_90
  ),
  
  Total_n = c(
    2*n_noninf_5pct_a20_min,
    2*n_noninf_5pct_a05_min,
    2*n_noninf_5pct_a20_80,
    2*n_noninf_5pct_a20_90,
    2*n_noninf_5pct_a05_80,
    2*n_noninf_5pct_a05_90,
    2*n_diff_5pct_a05_min,
    2*n_diff_5pct_a20_min,
    2*n_diff_5pct_a05_80,
    2*n_diff_5pct_a20_80,
    2*n_diff_15pct_a05_min,
    2*n_diff_15pct_a20_min,
    2*n_diff_15pct_a05_80,
    2*n_diff_15pct_a05_90,
    2*n_diff_15pct_a20_80,
    2*n_diff_15pct_a20_90
  ),
  
  Notes = c(
    "Minimum n (EFSA α)",
    "Minimum n (stringent α)",
    "EFSA recommended",
    "EFSA + high power",
    "Stringent α",
    "Stringent α + power",
    "Minimum n (conventional)",
    "Minimum n (lenient)",
    "Inappropriate test",
    "Inappropriate test",
    "Minimum n (conventional)",
    "Minimum n (lenient)",
    "Conventional approach",
    "Conventional + high power",
    "More lenient",
    "More lenient + power"
  )
)
```


```{r}

knitr::kable(comparison_table,booktabs=TRUE) ##|> kableExtra::kable_classic()
```


