---
title: "Manuscript Rvision"
format: html
editor: source
editor_options: 
  chunk_output_type: console
---



## To Dos:

- [x] Fix figure numbers and cross‑references; add explicit links from Figures 3–4 to Figure 1 (design/variance patterns).
- [x] Replace underscores in method names in running text (e.g., LM_Hetero → LM‑Hetero) while keeping underscores in code blocks to match R objects.
- [x] Spell out OECD on first mention and keep consistent thereafter.
- [x] Move the long Table 5 to SI; 
- [x] keep a streamlined summary table of Table 5 (now table s3) or narrative in the main text.
- [x] Replace underscores in Figures?
- [x] Reviewer 1 request, 4.2.3: In Section 4.2.3, start with “Influence of Experimental Design Parameters” (explicitly referencing Figure 1), then “Power to detect small effects,” then “Power to detect large effects.”
- [x] Add one sentence in Methods: you modeled treatment as fixed and tank as a random intercept in the individual‑level LMM, with individuals at the residual level; ICC is computed from variance components. Also note the caution about mislabeling replicate IDs across treatments and how that inflates Type I error.

## Regenerate the Figures

```{r}
sim4_decreasing_results <- readRDS("~/Projects/drcHelper/inst/RStats/sim4/decreasing_only/sim4_decreasing_results.rds")
test_methods <- list(
  "LMM_Homo" = lmm_dunnett_homo,
  "LMM_Hetero" = lmm_dunnett_hetero,
  "LM_Homo" = lm_dunnett_agg,
  "LM_Hetero" = gls_dunnett_agg,
  "Williams" = williams_test,
  "Jonckheere" = jonckheere_test,
  "Dunns" = manyone_dunns_test
)


# ICC values to explore
icc_values <- c(0.1, 0.3, 0.5, 0.7, 0.9)  # Low to high ICC

# Total variance (keeping this constant helps with interpretation)
total_variance <- 10

# Function to calculate variance components from ICC
calculate_variance_components <- function(icc, total_variance) {
  var_tank <- icc * total_variance
  var_individual <- (1 - icc) * total_variance
  return(list(var_tank = var_tank, var_individual = var_individual))
}

# Create variance component values for each ICC
variance_components <- lapply(icc_values, calculate_variance_components,
                              total_variance = total_variance)

# Extract as vectors for parameter grid
var_tank_values <- sapply(variance_components, function(x) x$var_tank)
var_individual_values <- sapply(variance_components, function(x) x$var_individual)

# Create named vectors for better readability in results
names(var_tank_values) <- paste0("ICC_", icc_values)
names(var_individual_values) <- paste0("ICC_", icc_values)

# Display the variance components
variance_table <- data.frame(
  ICC = icc_values,
  var_tank = var_tank_values,
  var_individual = var_individual_values,
  total_variance = total_variance
)
print(variance_table)


# Define parameter grid for full simulation
param_grid_1 <- expand.grid(
  n_doses = 5,
  min_dose = 0,
  max_dose = 20,
  m_tanks = c(4, 6),
  k_individuals = c(3,6,10),
  # var_tank = c(2, 4),
  # var_individual = c(4,6),
  max_effect = c(5, 20),
  response_type = c("decreasing"),
  alpha = 0.05,
  alternative = "less",
  stringsAsFactors = FALSE
)

param_grid_2 <- expand.grid(
  n_doses = 5,
  min_dose = 0,
  max_dose = 20,
  m_tanks = c(4, 6),
  k_individuals = c(3,6,10),
  # var_tank = c(2, 4),
  # var_individual = c(4,6),
  max_effect = c(20),
  response_type = c("none"),
  alpha = 0.05,
  alternative = "less",
  stringsAsFactors = FALSE
)

nrow(param_grid_1)
nrow(param_grid_2)
param_grid_1 <- rbind(param_grid_1,param_grid_2)
(36+6)*5 ## 180 scenarios, 4 non-control, ,
param_grid <- dplyr::cross_join(param_grid_1,variance_table)
dim(param_grid)
sim4_decreasing_results <-  dplyr::left_join(param_grid,sim4_decreasing_results)
sim4_decreasing_results_1 <- sim4_decreasing_results ##%>% dplyr::filter(!(response_type == "none"))
dim(sim4_decreasing_results_1)
```


```{r}
dim(sim4_decreasing_results_1)
sim4_full_results <- readRDS("~/Projects/drcHelper/inst/RStats/sim4/sim4_full_results.rds")
# Define test methods to compare
test_methods <- list(
  "LMM_Homo" = lmm_dunnett_homo,
  "LMM_Hetero" = lmm_dunnett_hetero,
  "LM_Homo" = lm_dunnett_agg,
  "LM_Hetero" = gls_dunnett_agg,
  "Williams" = williams_test,
  "Jonckheere" = jonckheere_test,
  "Dunns" = manyone_dunns_test
)


# ICC values to explore
icc_values <- c(0.1, 0.3, 0.5, 0.7, 0.9)  # Low to high ICC

# Total variance (keeping this constant helps with interpretation)
total_variance <- 10

# Function to calculate variance components from ICC
calculate_variance_components <- function(icc, total_variance) {
  var_tank <- icc * total_variance
  var_individual <- (1 - icc) * total_variance
  return(list(var_tank = var_tank, var_individual = var_individual))
}

# Create variance component values for each ICC
variance_components <- lapply(icc_values, calculate_variance_components,
                              total_variance = total_variance)

# Extract as vectors for parameter grid
var_tank_values <- sapply(variance_components, function(x) x$var_tank)
var_individual_values <- sapply(variance_components, function(x) x$var_individual)

# Create named vectors for better readability in results
names(var_tank_values) <- paste0("ICC_", icc_values)
names(var_individual_values) <- paste0("ICC_", icc_values)

# Display the variance components
variance_table <- data.frame(
  ICC = icc_values,
  var_tank = var_tank_values,
  var_individual = var_individual_values,
  total_variance = total_variance
)
print(variance_table)


# Define parameter grid for full simulation
param_grid_1 <- expand.grid(
  n_doses = 5,
  min_dose = 0,
  max_dose = 20,
  m_tanks = c(4, 6),
  k_individuals = c(3,6,10),
  # var_tank = c(2, 4),
  # var_individual = c(4,6),
  max_effect = c(5, 20),
  response_type = c("threshold","non_monotonic","oscillating"),
  alpha = 0.05,
  alternative = "less",
  stringsAsFactors = FALSE
)

param_grid_2 <- expand.grid(
  n_doses = 5,
  min_dose = 0,
  max_dose = 20,
  m_tanks = c(4, 6),
  k_individuals = c(3,6,10),
  # var_tank = c(2, 4),
  # var_individual = c(4,6),
  max_effect = c(5),
  response_type = c("none"),
  alpha = 0.05,
  alternative = "less",
  stringsAsFactors = FALSE
)

nrow(param_grid_1)
nrow(param_grid_2)
param_grid_1 <- rbind(param_grid_1,param_grid_2)
(36+6)*5 ## 180 scenarios, 4 non-control, ,
param_grid <- dplyr::cross_join(param_grid_1,variance_table)
dim(param_grid)

sim4_full_results <- dplyr::left_join(param_grid,sim4_full_results)
```



```{r}
full_results <- rbind(sim4_full_results,sim4_decreasing_results_1)
wide_res <- full_results %>% tidyr::pivot_wider(names_from = Method,values_from = Power)

wide_res %>% filter(Expected_Response ==100) %>% reframe(LMM_Homo=summary(LMM_Homo),
                                                         LMM_Hetero=summary(LMM_Hetero),
                                                         LM_Homo = summary(LM_Homo),
                                                         LM_Hetero = summary(LM_Hetero),
                                                         Jonckheere = summary(Jonckheere),
                                                         Williams = summary(Williams),
                                                         Dunns = summary(Dunns)
) %>% pander::pander(.)


wide_res %>% filter(Expected_Response ==95) %>% reframe(LMM_Homo=summary(LMM_Homo),
                                                        LMM_Hetero=summary(LMM_Hetero),
                                                        LM_Homo = summary(LM_Homo),
                                                        LM_Hetero = summary(LM_Hetero),
                                                        Jonckheere = summary(Jonckheere),
                                                        Williams = summary(Williams)
)

wide_res %>% filter(Expected_Response ==90) %>% reframe(LMM_Homo=summary(LMM_Homo),
                                                        LMM_Hetero=summary(LMM_Hetero),
                                                        LM_Homo = summary(LM_Homo),
                                                        LM_Hetero = summary(LM_Hetero),
                                                        Jonckheere = summary(Jonckheere),
                                                        Williams = summary(Williams)
)

out <- wide_res %>% group_by(Expected_Response) %>% nest() %>% mutate(power_summary=map(data,function(x){
  out <- data.frame(stat=c("Min","Q1","Mean","Median","Q3","Max"))
  out1 <- x %>% reframe(LMM_Homo=summary(LMM_Homo),
                        LMM_Hetero=summary(LMM_Hetero),
                        LM_Homo = summary(LM_Homo),
                        LM_Hetero = summary(LM_Hetero),
                        Jonckheere = summary(Jonckheere),
                        Williams = summary(Williams)
  )
  cbind(out,out1)
  
})) %>% dplyr::select(-data) %>% unnest(c(power_summary))

out %>% mutate(across(where(is.table),
                      as.numeric)) %>% knitr::kable(.,digits = 3) %>% kableExtra::kable_classic()
```

```{r}
out2 <- wide_res %>% filter(m_tanks ==4, k_individuals ==6, max_effect == 5) %>% dplyr::select(-c(n_doses,min_dose,max_dose,
                                                                                                  m_tanks,k_individuals,alpha,
                                                                                                  alternative,
                                                                                                  max_effect,
                                                                                                  total_variance))
out2 %>% filter(Expected_Response <100) %>% summarise(nJ=sum(Jonckheere>Williams),n=length(Jonckheere))%>% mutate(pnJ = nJ/n)
wide_res %>% filter(max_effect == 5) %>% filter(Expected_Response <100, max_effect == 5) %>% summarise(nJ=sum(Jonckheere>Williams),n=length(Jonckheere))%>% mutate(pnJ = nJ/n)
wide_res %>% filter(max_effect == 20) %>% filter(Expected_Response <100, max_effect == 20) %>% summarise(nJ=sum(Jonckheere>Williams),n=length(Jonckheere))%>% mutate(pnJ = nJ/n)
wide_res %>% filter(max_effect > 5) %>% filter(Expected_Response <100) %>% summarize(loss= max(Williams-Jonckheere))

wide_res %>% filter(max_effect > 5) %>% filter(Expected_Response <100, Dose_Level=="Dose_3") %>% summarize(loss= max(Williams-Jonckheere))
```


### Figures

```{r}
full_results <- rbind(sim4_full_results,sim4_decreasing_results_1)
full_results_1 <- full_results %>% dplyr::mutate(response_type = case_match(response_type,"decreasing"~"Decreasing","non_monotonic"~"Non-monotonic","none"~"No Effect","oscillating"~"Oscillating","threshold"~"Threshold")) %>% 
  mutate(Method = gsub("_","-",Method)) %>% mutate(Dose_Level= gsub("_","-",Dose_Level))
unique(full_results_1$response_type)
unique(full_results_1$Method)
unique(full_results_1$Dose_Level)
full_results <- full_results_1
design_effect <- full_results %>% group_by(max_effect,response_type,Dose_Level) %>%
  reframe(neffect=length(unique(Expected_Response)), Reduction = paste0(100-Expected_Response[1],"%"))
tempdata <- full_results
design_effect_temp <- design_effect
```


```{r}
# Theme with 12 pt text and 1 pt lines
publication_theme_pdf_12 <- function(base = theme_bw(), line_pt = 1.0) {
  base + theme(
    plot.title    = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 12),
    plot.caption  = element_text(size = 12),
    axis.title.x  = element_text(size = 12),
    axis.title.y  = element_text(size = 12),
    axis.text.x   = element_text(size = 12),
    axis.text.y   = element_text(size = 12),
    legend.title  = element_text(size = 12),
    legend.text   = element_text(size = 12),
    strip.text.x  = element_text(size = 12),
    strip.text.y  = element_text(size = 12),
    axis.line     = element_line(linewidth = line_pt / ggplot2::.pt),
    axis.ticks    = element_line(linewidth = line_pt / ggplot2::.pt),
    panel.grid    = element_line(linewidth = (line_pt * 0.75) / ggplot2::.pt)
  )
}

# Set defaults for geom text and lines
update_geom_defaults("text",    list(size = 12 / ggplot2::.pt))
update_geom_defaults("label",   list(size = 12 / ggplot2::.pt))
update_geom_defaults("line",    list(linewidth = 1.0 / ggplot2::.pt))
update_geom_defaults("segment", list(linewidth = 1.0 / ggplot2::.pt))
update_geom_defaults("smooth",  list(linewidth = 1.0 / ggplot2::.pt))
update_geom_defaults("point",   list(stroke    = 1.0 / ggplot2::.pt))

# theme_set(theme_bw(base_size = 18))
for(max_effect0 in c(5,20)){
  for(m_tank0 in c(4,6)){
    for(k_individuals0 in c(3,6,10)){
      
      ggplot(tempdata%>%
               dplyr::filter(m_tanks==m_tank0, k_individuals==k_individuals0,
                             max_effect==max_effect0),aes(x=ICC,y=Power,color=Method))+ geom_point(aes(pch=Method))+
        scale_shape_manual(values = 1:7)+
        facet_grid(response_type ~ Dose_Level,scales = "free")  +
        geom_line(alpha=0.5)+
        geom_hline(yintercept = c(0.05,0.8),lty=2,alpha=0.3)+ publication_theme_pdf_12()+
        theme(legend.position = "bottom")+ ## ggplot2::scale_color_viridis_d(direction = 1)
        ggthemes::scale_color_solarized()+ggtitle(paste0("4 tanks, ", k_individuals0, " individuals, maximum effect: ", max_effect0, "%"))+
        geom_text(data=design_effect_temp%>%dplyr::filter(max_effect==max_effect0), aes(x=0.5,y=1.01,label=Reduction),col= "black" ) 
      ##ggsave(paste0("SimPower_",m_tank0, "_tank_",k_individuals0,"_ind_",max_effect0,"_effect.png"),dpi=300, width = 6,height =6)
      ggsave(paste0(here::here(),"/inst/RStats/Revision/SimPower_",m_tank0, "_tank_",k_individuals0,"_ind_",max_effect0,"_effect.pdf"),dpi=300, width = 7.5,height = 9)
    }
  }
}## finish here
```


```{r}

```


## Heterogeneous Variance

```{r}
sim5_results <- readRDS("~/Projects/drcHelper/inst/RStats/sim5/archive/sim5_results.rds")

sim5_results %>% filter(response_type =="none") %>% filter(Method=="Jonckheere") -> tmp
sum(tmp$Power>0.05)/nrow(tmp)
sim5_results %>% filter(response_type =="none") %>% filter(Method=="Williams") -> tmp
sum(tmp$Power>0.05)/nrow(tmp)

# Variance pattern generator functions
variance_patterns <- list(
  homogeneous = function(n_doses, base_var = 4) {
    rep(base_var, n_doses)
  },
  
  increasing = function(n_doses, base_var = 4) {
    base_var * seq(0.5, 2, length.out = n_doses)
  },
  
  decreasing = function(n_doses, base_var = 4) {
    base_var * seq(2, 0.5, length.out = n_doses)
  },
  
  v_shaped = function(n_doses, base_var = 4) {
    mid_point <- ceiling(n_doses/2)
    c(
      base_var * seq(2, 1, length.out = mid_point),
      base_var * seq(1, 2, length.out = n_doses - mid_point + 1)[-1]
    )
  }
)

tank_level_methods <- list(
  "LM_Homo" = lm_dunnett_agg_simple,
  "LM_Hetero" = gls_dunnett_agg_simple,
  "Williams" = williams_test,
  "Jonckheere" = jonckheere_test,
  "Dunns" = manyone_dunns_test
)

# Visualize the variance patterns
# variance_plot <- plot_variance_patterns(n_doses = 5, base_variance = 2)
# print(variance_plot)

# Parameter grid for heterogeneous variance question
param_grid_hetero <- expand.grid(
  n_doses = 5,
  min_dose = 0,
  max_dose = 20,
  m_tanks = c(4,6,10),
  variance_pattern = c("homogeneous", "increasing", "decreasing", "v_shaped"),
  base_variance = c(4, 6),
  max_effect = c(5, 20),  # Null and moderate effect
  response_type = c("none", "threshold", "oscillating"),  # Focus on null and decreasing
  alpha = 0.05,
  alternative = "less",
  stringsAsFactors = FALSE
)

nrow(param_grid_hetero)
## 120 ## 24*2*3
param_grid_hetero <- param_grid_hetero %>% dplyr::filter(max_effect == 5 | (max_effect == 20 & response_type !="none"))

sim5_results$Method[sim5_results$Method=="GLS_Hetero"] <- "LM_Hetero"
sim5_results$Method[sim5_results$Method=="LM_Hetero"] <- "LM-Hetero"
sim5_results$Method[sim5_results$Method=="LM_Homo"] <- "LM-Homo"
base_variance_names <- c("4" = "Base variance: 4",
                         "6" = "Base variance: 6")
sim5_results$variance_pattern <- gsub("_","-",sim5_results$variance_pattern )
sim5_results$Dose_Level <- gsub("Dose_","Dose-",paste0(sim5_results$Dose_Level,"\n",(100-sim5_results$Expected_Response),"%"))

sim5_results <- sim5_results %>% dplyr::mutate(response_type = case_match(response_type,"decreasing"~"Decreasing","non_monotonic"~"Non-monotonic","none"~"No Effect","oscillating"~"Oscillating","threshold"~"Threshold"))
design_effect <- sim5_results %>% group_by(max_effect,response_type,Dose_Level,variance_pattern,base_variance) %>%
  reframe(neffect=length(unique(Expected_Response)), Reduction = paste0(100-Expected_Response[1],"%"),
          nvar=length(unique(Variance)),Variance=Variance[1])

tempdata <- sim5_results

for(max_effect0 in c(5,20)){
  for(m_tank0 in c(4,6,10)){
    for(response_type0 in c("No Effect","Oscillating","Threshold")){
      if(max_effect0 == 20 && response_type0 =="No Effect"){}else{
        ggplot(tempdata%>%
                 dplyr::filter(m_tanks==m_tank0, response_type == response_type0,
                               max_effect==max_effect0)%>%droplevels(.),
               aes(x=Dose_Level,y=Power,color=Method))+
          geom_point(aes(pch=Method))+
          # facet_grid( variance_pattern~base_variance,scales = "free")  +
          facet_grid( variance_pattern~base_variance,scales = "free",labeller = labeller(base_variance=base_variance_names)) +
          #geom_line()+
          geom_hline(yintercept = c(0.05,0.8),lty=2,alpha=0.3)+
          scale_y_continuous(breaks = seq(0, 1, by = 0.2),limits = c(0, 1.15)) + publication_theme_pdf_12()+
          theme(legend.position = "bottom")+ ## ggplot2::scale_color_viridis_d(direction = 1)
          ggthemes::scale_color_solarized()+ggtitle(paste0("Response type: ", response_type0,", ", m_tank0, " tanks, ", "maximum effect: ", max_effect0, "%"))+
          geom_text(data=design_effect%>%dplyr::filter(max_effect==max_effect0,response_type == response_type0)%>%droplevels(.),
                    aes(x=Dose_Level,y=1.1,label=paste("Var=",Variance)), col= "black" ) + xlab("Dose Level")
        #ggsave(paste0(here::here(),"/inst/RStats/Revision/SimHetero_",m_tank0, "_tank_","response_", response_type0,max_effect0,"_effect.png"),
        #dpi=300, width = 6,height = 6)
        ggsave(paste0(here::here(),"/inst/RStats/Revision/SimHetero_",m_tank0, "_tank_","response_", response_type0,max_effect0,"_effect.pdf"),
               dpi=300, width = 7,height = 7)
        
      }
      
      
    }
  }
}


for(max_effect0 in c(5,20)){
  for(m_tank0 in c(4,6,10)){
    for(response_type0 in c("No Effect")){
      if(max_effect0 == 20 && response_type0 =="No Effect"){}else{
        ggplot(tempdata%>%
                 dplyr::filter(m_tanks==m_tank0, response_type == response_type0,
                               max_effect==max_effect0)%>%droplevels(.),
               aes(x=Dose_Level,y=Power,color=Method))+
          geom_point(aes(pch=Method))+
          # facet_grid( variance_pattern~base_variance,scales = "free")  +
          facet_grid( variance_pattern~base_variance,scales = "free",labeller = labeller(base_variance=base_variance_names)) +
          #geom_line()+
          geom_hline(yintercept = c(0.05,0.8),lty=2,alpha=0.3)+
          scale_y_continuous(breaks = seq(0, 1, by = 0.2),limits = c(0, 0.12)) + publication_theme_pdf_12()+
          theme(legend.position = "bottom")+ ## ggplot2::scale_color_viridis_d(direction = 1)
          ggthemes::scale_color_solarized()+ggtitle(paste0("Response type: ", response_type0,", ", m_tank0, " tanks, ", "maximum effect: ", max_effect0, "%"))+
          geom_text(data=design_effect%>%dplyr::filter(max_effect==max_effect0,response_type == response_type0)%>%droplevels(.),
                    aes(x=Dose_Level,y=0.1,label=paste("Var=",Variance)), col= "black" ) + xlab("Dose Level")
        ggsave(paste0(here::here(),"/inst/RStats/Revision/SimHetero_",m_tank0, "_tank_","response_", response_type0,max_effect0,"_effect.png"),dpi=300, width = 7.5,height = 7)
        ggsave(paste0(here::here(),"/inst/RStats/Revision/SimHetero_",m_tank0, "_tank_","response_", response_type0,max_effect0,"_effect.pdf"),
               dpi=300, width = 7,height = 7)
        
      }
      
      
    }
  }
}

```

## References that are not cited in revised manuscript but could be useful

- Shan, G., Young, D., and Kang, L. (2014) A new powerful nonparametric rank test for ordered alternative problem. PLoS One https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0112924

- Ritz  C. (2010) Toward a unified approach to dose-response modeling in ecotoxicology. Environ Toxicol Chem, 29(1):220-9. https://doi.org/10.1002/etc.7. PMID: 20821438.
- Bruce , R..D. and Versteeg,, D.J.(1992) A Statistical Procedure for Modelling Continuous Ecotoxicity Data. Environ. Toxicol. Chem.11, 1485-1494 

- Munzel. U., Hothorn, L.A. (2001). A unified Approach to Simultaneous Rank Tests Procedures in the Unbalanced One-way Layout. Biometric Journal, 43, 553-569.

- Zeileis, A., Köll, S., & Graham, N. (2020). Various Versatile Variances: An Object-Oriented Implementation of Clustered Covariances in R. Journal of Statistical Software, 95(1), 1–36. https://doi.org/10.18637/jss.v095.i01


- Hothorn, T., F. Bretz, P.H.Westfall, R. M. Heiberger, A. Schuetzenmeister, and S. Scheibe. (2021). multcomp: Simultaneous Inference in General Parametric Models. R Package Version 1.4-16. http://CRAN.R-project.org/package=multcomp.
- Hothorn, T., Bretz, F, Westfall, P. (2008). “Simultaneous Inference in General Parametric Models.” Biometrical Journal, 50(3), 346-363.
- Hothorn, T., F. Bretz, P.H.Westfall, R. M. Heiberger, A. Schuetzenmeister, and S. Scheibe. (2021). multcomp: Simultaneous Inference in General Parametric Models. R Package Version 1.4-16. http://CRAN.R-project.org/package=multcomp.

- Konietschke, F., Placzek, M., Schaarschmidt F., Hothorn, L.A. (2015). nparcomp: An R Software  Package for Nonparametric Multiple Comparisons and Simultaneous Confidence Intervals. Journal of Statistical Software, 64(9), 1-17. URL http://www.jstatsoft.org/v64/i09/. 

- Fox D.R. (2010). A Bayesian approach for determining the no effect concentration and hazardous concentration in ecotoxicology. Ecotoxicol Environ Saf. 73(2):123-31. doi: 10.1016/j.ecoenv.2009.09.012. PMID: 19836077.

## CPCAT

The reality of ecotoxicological count data, where over-dispersion, under-dispersion, non-independence, inhomogeneous variance without a pattern, and small sample size are common challenges. 

## Equivalence Testing

quote for the revised EFSA bee guidance document: 

> The statistical analysis should be conducted according to the general approach for higher tier studies (see Annex C of the guidance document and Chapter 10.1). The applicant should apply a one-sided equivalence test (α = 0.2) for each endpoint, with an equivalence limit corresponding to a 10% reduction in the treated test compared to the control, to prove that there are no adverse effects on foraging behaviour. 

This, in our opinion, is to cut the feet to fit the shoes. Just to use a seemingly better approach, without considering is the method switch requires complete experiment re-design and whether the scientific goal is verifiable. In fact, the regulatory goal may be empirically unverifiable with the proposed framework under standard use of alpha.


## Trend Tests

The redundancy arises because a **trend test** (like the Jonckheere-Terpstra test or Williams' test) is specifically designed to test the null hypothesis of *no trend* versus the alternative hypothesis of a **monotonic trend** (e.g., increasing or decreasing response with dose).

### 1. The Trend Test Implies Monotonicity

The alternative hypothesis of a one-sided **trend test** directly assumes the response is either monotonically increasing or monotonically decreasing across dose groups.

* If you **reject the null hypothesis** of *no difference*, you are accepting the alternative: there is a significant monotonic trend.
* If you **fail to reject the null hypothesis**, it suggests there's no sufficient evidence of the monotonic trend the test is looking for.

### 2. What does the monotonicity test do?

A separate "monotonicity test" only tells you *if the assumption is plausible*, but it doesn't perform the final test of interest (i.e., whether the trend is statistically significant). The trend test is the ultimate goal.

### 3. The Williams' Test Example

For tests like the **Williams' test**, the procedure is even more explicitly tied to monotonicity: it uses the **Pool Adjacent Violators Algorithm (PAVA)** to impose a monotonic fit on the data before calculating the statistic. This means the test is *already enforcing* the assumption it needs, regardless of what a pre-test might say. The test statistic is calculated specifically to look for the best monotonic fit.

### Why Flowcharts Include Monotonicity Test before a Trend Test (The Justification?)

Despite the logical redundancy, flowcharts sometimes include a separate check for monotonicity for pedagogical or overly cautious reasons:

1.  **Guiding Test Selection:** It serves as a clear decision point to distinguish between **trend tests** (monotonic alternatives) and **any-difference tests** (non-monotonic alternatives, like ANOVA or Dunnett's test for many-to-one comparison).
2.  **Addressing Model Validity:** Some researchers want to ensure that if they run a powerful trend test, the data *visually and statistically* support the underlying structure (monotonicity) before applying the trend test, especially if the data are non-normal or highly variable.

## Other Thoughts

"削足适履" translates to "to cut the feet to fit the shoes" and describes a situation where you force something to fit a preconceived or unsuitable mold, often resulting in an impractical or inelegant solution. It can also be expressed as "to act in a Procrustean manner" or, more generally, as a situation where you "force something to fit". 

- Literal meaning: To cut one's feet to make them fit a shoe.
- Figurative meaning: Forcing a situation to conform to a rigid or inappropriate standard, despite the negative consequences.
- English equivalents:
- To cut the feet to fit the shoes
- To act in a Procrustean manner
- To force something to fit
- An impractical or inelegant solution 

## Goal of the Paper

### Questions to answer:

-	Under what conditions can standard decision flowcharts recommend analyses that yield misleading NOEC inferences, and how should analysts respond when visual patterns contradict flowchart‑prescribed tests
- Power vs. design, practical constraints, reliability and intuitiveness, and the scientific goal. 


