---
title: "drcHelper"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{drcHelper}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---



```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",  
  warning = FALSE,
  message = FALSE,
  fig.width = 7
)
library(dplyr)
library(ggplot2)
## library(tidyverse)
```

```{r setup}
library(drcHelper)

```

In this vignette, we demonstrate how to use the **{drcHelper}** package to create tables and graphs for inclusion in a dose-response analysis report. While the functions were developed some time ago and may not represent the most elegant solutions, they are functional. We welcome any suggestions for improvement. 

## Preliminary Summary

```{r}
data("dat_medium")
dat_medium <- dat_medium %>% mutate(Treatment=factor(Dose,levels=unique(Dose))) 
dat_medium$Response[dat_medium$Response < 0] <- 0
prelimPlot3(dat_medium)
prelimSummary(dat_medium) %>% knitr::kable(.,digits = 3)
```

## Fitting multiple models and rank them.

```{r}
mod <- drm(Response~Dose,data=dat_medium,fct=LL.3())
fctList <- list(LN.4(),LL.4(),W1.3(),LL2.2())
# plot(mod,type="all")
res <- mselect.plus(mod,fctList = fctList )
modList <- res$modList
res$Comparison

drcCompare(modRes=res)
```


```{r}
library(purrr)
edResTab <- mselect.ED(modList = modList,respLev = c(10,20,50),trend="Decrease",CI="inv")
edResTab
```


## Plot multiple models together

```{r}
p <- plot.modList(modList[1:3])
p
```


## Adding ECx and ECx CI's to the plots

```{r}
p1 <- plot.modList(modList[1])
addECxCI(p1,object=modList[[1]],EDres=NULL,trend="Decrease",endpoint="EC", respLev=c(10,20,50),
                     textAjust.x=0.01,textAjust.y=0.3,useObsCtr=FALSE,d0=NULL,textsize = 4,lineheight = 0.5,xmin=0.012)+ ylab("Response Variable [unit]") + xlab("Concentration [µg a.s./L]")
## addECxCI(p)
```

## Report ECx

```{r}
resED <- t(edResTab[1:3, c(2,4,5,6)])
colnames(resED) <- paste("EC", c(10,20,50))
knitr::kable(resED,caption = "Response Variable at day N",digits = 3)
```
**Calculate specific ECx:  **
```{r}
mod <-modList[[1]]
edres <- ED.plus(mod,c(5,10,20,50),trend="Decrease")
edres%>%knitr::kable(.,digits = 3)
```


## Model Output

```{r}
modsum <- summary(mod)
knitr::kable(coef(modsum),digits = 3)
```







## Additional Notes

_To be written_

### Dependencies

```r
library(drcHelper)
library(DependenciesGraphs)
dep <- funDependencies("package:drcHelper","ED.plus")
plot(dep)
dep <- funDependencies("package:drcHelper","mselect.ED")
plot(dep)

#dep <- funDependencies("package:drcHelper","mselect.plus")
#plot(dep)

dep <- envirDependencies('package:drcHelper')
plot(dep)
```
