---
title: "Dunnett's Test for Data with Hierarchical Structure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dunnetts_Test_for_Data_with_Hierarchical_Structure}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7
)
```

```{r setup}
library(drcHelper)
library(nlme)
library(lme4)
library(multcomp)
```

**drcHelper** provides a wrapper function `dunnett_test` to use the `multcomp` package to perform Dunnett tests on both the mixed model for individual data (`ind_model`) and the tank-level linear model for aggregated data (`tank_model`).

First we simulate a dataset with log-logistic dose-response. 

```{r}
# Let's simulate a single dataset and examine its properties
set.seed(123)

# Simulate dose-response data with specific variance components
sim_data <- simulate_dose_response(
  n_doses = 5,
  dose_range = c(0, 20),
  m_tanks = 4,
  k_individuals = 10,
  var_tank = 6,     # Between-tank variance
  var_individual = 2,  # Within-tank (individual) variance
  include_individuals = TRUE,
  response_function = function(dose) {
    # Simple linear dose-response with threshold at dose 10
    ifelse(dose > 10, 5 + 2 * (dose - 10), 5)
  }
)

# Calculate theoretical ICC
theoretical_icc <- 6 / (6 + 2)  # var_tank / (var_tank + var_individual)
cat("Theoretical ICC:", theoretical_icc, "\n")

# Examine the data structure
head(sim_data)
str(sim_data)

# Visualize the data to see the hierarchical structure
library(ggplot2)

# Plot individual data points with tank means
ggplot(sim_data, aes(x = factor(Dose), y = Response, color = factor(Tank))) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun = mean, geom = "line", aes(group = factor(Tank))) +
  labs(title = "Dose-Response Data with Hierarchical Structure",
       x = "Dose", y = "Response",
       color = "Tank") +
  theme_minimal()

# Calculate observed ICC using a mixed model
library(lme4)
mixed_model <- lmer(Response ~ factor(Dose) + (1|Tank), data = sim_data)
summary(mixed_model)

# Extract variance components
vc <- VarCorr(mixed_model)
tank_var <- as.numeric(vc$Tank)
residual_var <- attr(vc, "sc")^2
observed_icc <- tank_var / (tank_var + residual_var)
cat("Observed ICC:", observed_icc, "\n")

# Aggregate data to tank level
tank_data <- aggregate(Response ~ Dose + Tank, data = sim_data, FUN = mean)
head(tank_data)

# Compare individual-level model with tank-level model
# Individual level (mixed model)
ind_model <- lmer(Response ~ factor(Dose) + (1|Tank), data = sim_data)

# Tank level (regular linear model)
tank_model <- lm(Response ~ factor(Dose), data = tank_data)

# Compare model summaries
summary(ind_model)
summary(tank_model)

# Extract and compare fixed effects
ind_fixed <- fixef(ind_model)
tank_fixed <- coef(tank_model)
```


## Perform Dunnett Test for Different type of Models


### Homoscedastic mixed model 

```{r}
# 1. Homoscedastic mixed model (equivalent to ind_model1)
sim_data$Treatment <- factor(sim_data$Dose)
result1 <- dunnett_test(
  data = sim_data,
  response_var = "Response",
  dose_var = "Treatment",  # Using your Treatment factor
  tank_var = "Tank",
  include_random_effect = TRUE,
  variance_structure = "homoscedastic"
)
result1
```

This is equivalent to 

```{r}
ind_model1 <- lmer(Response ~ Treatment + (1 | Tank),sim_data) ## homoscedastic errors
# Apply Dunnett test to mixed model
ind_dunnett1 <- glht(ind_model1, linfct = mcp(Treatment = "Dunnett"))

s1 <- summary(ind_dunnett1)
result1$results_table$p.value - s1$test$pvalues < 1e-04
```

### Heteroscedastic mixed model

```{r}
# 2. Heteroscedastic mixed model (equivalent to  mod.nlme)
result2 <- dunnett_test(
  data = sim_data,
  response_var = "Response",
  dose_var = "Treatment",
  tank_var = "Tank",
  include_random_effect = FALSE,
  variance_structure = "heteroscedastic"
)


result2
```




```{r}
mod.nlme <- nlme::lme(Response ~ Treatment ,  
                       random  = ~ 1 | Tank, 
                       weights = varIdent(form= ~1|Treatment),
                       data    = sim_data)
ind_dunnett2 <- glht(mod.nlme, linfct = mcp(Treatment = "Dunnett"))
s2 <- summary(ind_dunnett2)
result2$results_table$p.value - s2$test$pvalues < 1e-05
result2$results_table$estimate -s2$test$coefficients
```



### Homoscedastic fixed effect model 


```{r}
# Tank level (regular linear model)
tank_data$Treatment <- factor(tank_data$Dose)
result3 <- dunnett_test(
  data = tank_data,
  response_var = "Response",
  dose_var = "Treatment",
  tank_var = "Tank",
  include_random_effect = FALSE,
  variance_structure = "homoscedastic"
)
result3
```

This is equivalent to:

```{r}
# Tank level (regular linear model)
tank_model <- lm(Response ~ Treatment, data = tank_data)
summary(glht(tank_model,linfct = mcp(Treatment="Dunnett")))
```

### Heteroscedastic fixed effect model 
```{r}
result4 <- dunnett_test(
  data = tank_data,
  response_var = "Response",
  dose_var = "Treatment",
  tank_var = NULL,
  include_random_effect = FALSE,
  variance_structure = "heteroscedastic"
)
result4
```

This is equivalent to: 

```{r}
library(nlme)
gls0 <- gls(Response ~ Treatment, data=tank_data,weights=varIdent(form= ~1|Treatment))
ind_gls0 <- glht(gls0, linfct = mcp(Treatment = "Dunnett"))

s4 <- summary(ind_gls0)
result4$results_table$estimate -s4$test$coefficients
result4$results_table$p.value - s4$test$pvalues < 1e-05
```

## Dunnett's Test with Different R Packages

We recommend to use the `multcomp` package to perform the routine Dunnett's test. 

```{r}
# Define Methods and Run All
methods_to_test <- c("Dunnett_PMCMRplus", "Dunnett_DescTools", "Dunnett_multcomp")

all_results <- purrr::map_dfr(methods_to_test, ~ broom_dunnett(
  x = Response ~ factor(Dose), 
  data = dat_medium,
  method = .x, 
  alternative = "two.sided"
))

all_results %>% knitr::kable(.,digits = 3)
```

