---
title: "Example_Analysis_Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example_Analysis_Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(drcHelper)
```


For a particular endpoint, analysis should follow the recommended flowchart in OECD 54 or in specific test guidelines. Deviations froom a standard procedure should be justified.


In this vignette, we are going to use dataset oecd201 as an example.

## Data overview


## Assumptions check

Pretests or checks could be conducted before testing, however, assumptions could also be checked after modelling, using a more holistic approach. Here we follow the routine procedure. 

> To allow the choice of the appropriate statistical tests, the normality and homoscedasticity (variance homogeneity) of a certain endpoint (e.g., shoot dry weight) data were checked.



### Check for normality


## NOER determination


## ER and ER determination 



## An overview of the endpoints

```{r}

```


The NOER, LOER, ER25 and ER50 for survival, plant height and shoot dry weight expressed as g a.s./ha
are summarized for each of the plant species for the final assessment (21 days after application) and can
be found in the following tables.


