---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This package includes a collection of helping functions for dose-response analysis that are generic and might be useful outside our own organization.

There are lots of examples for doing various type of dose-response analysis. Pros and cons for different methods are illustrated. Verification or validation of certain functions from certain packages are also included. 


```{r setup}
library(drcHelper)
```




## Other useful packages 

There are several packages I find useful for the statistical analysis of (eco)toxicological data.

### ecotoxicology package

 

- https://cran.r-project.org/web/packages/ecotoxicology/index.html

Implementation of the EPA's Ecological Exposure Research Division (EERD) tools (discontinued in 1999) for Probit and Trimmed Spearman-Karber Analysis. Probit and Spearman-Karber methods from Finney's book "Probit analysis a statistical treatment of the sigmoid response curve" with options for most accurate results or identical results to the book. Probit and all the tables from Finney's book (code-generated, not copied) with the generating functions included. Control correction: Abbott, Schneider-Orelli, Henderson-Tilton, Sun-Shepard. Toxicity scales: Horsfall-Barratt, Archer, Gauhl-Stover, Fullerton-Olsen, etc.

### NSEC package


