---
title: "Statistical Test Validation Framework - probit"
author: "Automated Validation System"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(drcHelper)
library(kableExtra)
library(ggplot2)

# Load test framework configuration
source("../config/test_framework_config.R")
```

# Probit Analysis Validation Report

## Executive Summary

This document presents comprehensive validation results for the **Probit Analysis** implementation against V-COP expected results. The validation covers:

- **Function Groups**: FG00430, FG00435
- **Test Alternatives**: N/A
- **Key Metrics**: Log10 (rate), Uncorrected, Corrected, Intercept, Slope

```{r load_data, results='asis'}
# Load test cases data
data("test_cases_data")
data("test_cases_res")

cat("**Dataset dimensions:**\n\n")
cat("- Test cases data: ", nrow(test_cases_data), " rows, ", ncol(test_cases_data), " columns\n")
cat("- Expected results: ", nrow(test_cases_res), " rows, ", ncol(test_cases_res), " columns\n\n")
```

## Test Configuration

```{r test_config, results='asis'}
# Define test configuration
TEST_NAME <- "probit"
FUNCTION_GROUPS <- get_function_groups(TEST_NAME)
TEST_CONFIG <- STATISTICAL_TESTS[[TEST_NAME]]

cat("**Test Configuration:**\n\n")
cat("- **Test Name:** ", TEST_CONFIG$name, "\n")
cat("- **Function Groups:** ", paste(FUNCTION_GROUPS, collapse = ", "), "\n")
cat("- **Test Function:** ", TEST_CONFIG$test_function, "\n")
cat("- **Implemented:** ", ifelse(TEST_CONFIG$implemented, "✅ Yes", "⚠️ No"), "\n\n")

if(!TEST_CONFIG$implemented) {
  cat("> ⚠️ **WARNING:** This test is not yet implemented. This template shows the validation framework structure.\n\n")
}
```

## Data Preparation and Validation

```{r data_preparation, results='asis'}
# Filter expected results for this test's function groups
expected_results <- test_cases_res[test_cases_res[['Function group ID']] %in% FUNCTION_GROUPS, ]

cat("**Expected results for ", TEST_CONFIG$name, ":**\n\n")
cat("- **Total expected results:** ", nrow(expected_results), "\n")
cat("- **Unique studies:** ", length(unique(expected_results[['Study ID']])), "\n\n")

# Show breakdown by function group
cat("**Breakdown by Function Group:**\n\n")
fg_summary <- table(expected_results[['Function group ID']])
for(i in seq_along(fg_summary)) {
  cat("- ", names(fg_summary)[i], ": ", fg_summary[i], " test cases\n")
}
cat("\n")
```

## Validation Methodology

The validation process follows these steps:

1. **Data Matching**: Match test case data with expected results by Study ID
2. **Test Execution**: Run Probit Analysis with appropriate parameters
3. **Result Comparison**: Compare actual vs expected values with tolerance-based validation
4. **Statistical Summary**: Aggregate validation results and success rates

```{r validation_framework}
# Validation function framework
run_probit_validation <- function(study_ids = NULL, alternatives = NULL) {
  
  if(is.null(study_ids)) {
    study_ids <- unique(expected_results[['Study ID']])
  }
  
  if(is.null(alternatives)) {
    alternatives <- if(!is.null(TEST_CONFIG$alternatives)) TEST_CONFIG$alternatives else c("two.sided")
  }
  
  validation_results <- list()
  
  for(study_id in study_ids) {
    cat("Processing study:", study_id, "\n")
    
    # Get test data for this study
    study_data <- test_cases_data[test_cases_data[['Study ID']] == study_id, ]
    
    if(nrow(study_data) == 0) {
      cat("  No test data found for study", study_id, "\n")
      next
    }
    
    # Get expected results for this study  
    study_expected <- expected_results[expected_results[['Study ID']] == study_id, ]
    
    if(nrow(study_expected) == 0) {
      cat("  No expected results found for study", study_id, "\n")
      next
    }
    
    for(alt in alternatives) {
      test_name <- paste(study_id, alt, sep = "_")
      
      validation_results[[test_name]] <- list(
        study_id = study_id,
        alternative = alt,
        test = test_name,
        passed = FALSE,  # Will be updated when test is implemented
        time = 0,
        details = list(
          note = "Test not yet implemented - framework structure only",
          n_comparisons = nrow(study_expected),
          n_passed = 0
        )
      )
      
      # TODO: Implement actual test execution when test function is available
      # if(TEST_CONFIG$implemented) {
      #   result <- do.call(TEST_CONFIG$test_function, list(
      #     data = study_data,
      #     alternative = alt,
      #     # Add other parameters as needed
      #   ))
      #   
      #   # Validate results against expected values
      #   # validation_results[[test_name]] <- validate_test_results(result, study_expected, alt)
      # }
    }
  }
  
  return(validation_results)
}

# Basic functionality tests framework  
basic_functionality_tests <- function() {
  
  basic_tests <- list()
  
  # Test 1: Basic function execution
  if(TEST_CONFIG$implemented) {
    # TODO: Add real basic functionality tests when implemented
    basic_tests[["Basic Function Execution"]] <- list(
      test = "Basic Function Execution",
      passed = FALSE,
      time = 0,
      details = "Test function not yet implemented"
    )
  } else {
    basic_tests[["Framework Structure"]] <- list(
      test = "Framework Structure",
      passed = TRUE,
      time = 0.001,
      details = "Validation framework structure verified"
    )
  }
  
  return(basic_tests)
}
```

## Test Execution

```{r execute_tests, results='asis'}
if(TEST_CONFIG$implemented) {
  cat("**Executing validation tests...**\n\n")
  
  # Run validation tests
  test_results <- run_probit_validation()
  
  # Run basic functionality tests
  basic_tests <- basic_functionality_tests()
  
  cat("✅ **Validation completed.**\n\n")
} else {
  cat("> ℹ️ **Note:** Test implementation not available - showing framework structure only.\n\n")
  
  # Create placeholder results to demonstrate framework
  test_results <- list(
    "PLACEHOLDER_less" = list(
      study_id = "PLACEHOLDER",
      alternative = "less", 
      test = "PLACEHOLDER_less",
      passed = FALSE,
      time = 0,
      details = list(note = "Placeholder - awaiting implementation")
    )
  )
  
  basic_tests <- basic_functionality_tests()
}
```

## Results Summary

```{r results_summary}
# Convert test results to summary format
validation_tests_list <- list()
for(test_name in names(test_results)) {
  validation_tests_list[[test_name]] <- list(
    test = test_name,
    passed = test_results[[test_name]]$passed,
    time = test_results[[test_name]]$time
  )
}

all_results <- c(validation_tests_list, basic_tests)

# Create summary table
test_summary <- data.frame(
  Test = sapply(all_results, function(x) x$test),
  Status = sapply(all_results, function(x) ifelse(x$passed, "✅ PASS", "❌ FAIL")),
  Time = sapply(all_results, function(x) sprintf("%.3f sec", x$time)),
  stringsAsFactors = FALSE
)

# Display results
kable(test_summary) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  row_spec(which(grepl("❌ FAIL", test_summary$Status)), background = "#FFCCCC") %>%
  row_spec(which(grepl("✅ PASS", test_summary$Status)), background = "#CCFFCC")

cat("Total Tests:", nrow(test_summary), "\n")
cat("Passed:", sum(grepl("✅ PASS", test_summary$Status)), "\n")
cat("Failed:", sum(grepl("❌ FAIL", test_summary$Status)), "\n")
cat("Success Rate:", round(100 * sum(grepl("✅ PASS", test_summary$Status)) / nrow(test_summary), 1), "%\n")
```

## Implementation Status

```{r implementation_status}
if(!TEST_CONFIG$implemented) {
  cat("📋 IMPLEMENTATION REQUIRED:\n\n")
  cat("To complete this validation, the following components need to be implemented:\n\n")
  cat("1. **Test Function**: ", TEST_CONFIG$test_function, "\n")
  cat("   - Input: test data, alternative hypothesis, other parameters\n")
  cat("   - Output: results structure with key metrics\n\n")
  cat("2. **Key Metrics Extraction**:\n")
  for(metric in TEST_CONFIG$key_metrics) {
    cat("   -", metric, "\n")
  }
  cat("\n3. **Alternative Hypothesis Support**:\n")
  if(!is.null(TEST_CONFIG$alternatives)) {
    for(alt in TEST_CONFIG$alternatives) {
      cat("   -", alt, "\n")
    }
  } else {
    cat("   - Not applicable (single test type)\n")
  }
  cat("\n4. **Integration with Validation Framework**:\n")
  cat("   - Update run_", TEST_NAME, "_validation() function\n")
  cat("   - Add result validation logic\n")
  cat("   - Implement basic functionality tests\n")
} else {
  cat("✅ Implementation completed - validation results above show actual test performance.\n")
}
```

## Visualization

```{r visualization}
if(nrow(test_summary) > 0) {
  # Create visualization
  test_summary$Time_Numeric <- as.numeric(gsub(" sec", "", test_summary$Time))
  test_summary$Status_Clean <- ifelse(grepl("✅ PASS", test_summary$Status), "PASS", "FAIL")
  
  ggplot(test_summary, aes(x = reorder(Test, Time_Numeric), y = Time_Numeric, fill = Status_Clean)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    labs(title = "Probit Analysis - Test Execution Time", 
         x = "Test Case", 
         y = "Time (seconds)") +
    scale_fill_manual(values = c("PASS" = "darkgreen", "FAIL" = "red")) +
    theme_minimal() +
    theme(axis.text.y = element_text(size = 8))
}
```

## Conclusion

This validation framework provides the structure for comprehensive Probit Analysis validation. The test implementation is pending. This framework provides the structure for validation once the test function is implemented.

### Next Steps

1. Implement
probit_test
function
2. Add result validation logic
3. Implement basic functionality tests
4. Run full validation suite

---

**Generated on:** `r Sys.time()`  
**Framework Version:** 1.0  
**Test Status:** PENDING IMPLEMENTATION
