---
format:
  docx:
    toc: true
    colorlinks: true
    fig-width: 7
    fig-height: 7
execute: 
  echo: true
knitr:
  opts_chunk:
    fig.path: figures/
    warning: false
    message: false
date: "`r Sys.Date()`"
date-format: long
bibliography: references.bib
---

## Definition of Test Cases

In addition to these test cases, we are also verifying if the selected R functions for a particular purpose. The functions, data, and expected results are listed in 



Test cases and expected results are stored in the excel file "R-V-Cop_test_cases_level_2_gesamt_20240911.xlsx" in the raw-data folder. The compound or study related information are redacted and it will not affect the numerical evaluation or verification. 

The data included in the test cases are either real data excluding the compound or chemical information or synthetic data generated to cover a special case. Most of the data are repeatedly used in the verification or validation of different tools for the statistical analyses of (eco)toxicology data. 

[@Box1976]


```{r setup0, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, results = "asis")
library(dplyr)

SummariseResults <- function(res) {
  res %>% summarise_if(is.numeric, sum) %>% knitr::kable()
}
WriteResults <- function(res) {
  purrr::pwalk(list(res$file, res$Categories), WriteFileResults)
}
WriteFileResults <- function(file, categories, ...) {
  cat("###", file, "\n")
  purrr::pwalk(list(categories$context, categories$Results), WriteCategoryResults)
}
WriteCategoryResults <- function(category, tests, ...) {
  cat("####", category, "\n")
  print(knitr::kable(tests))
}

path0 <- here::here()
devtools::load_all(paste0(path0))
```

```{r GenerateResults,results='hide',message=FALSE,warning=FALSE,fig.keep='none',include=FALSE}

file_lists <- c("ST_TestCase3.1.R","ST_TestCase3.2.R","ST_TestCase3.3.R")
file_lists <- list.files(paste0("path0","/inst/SystemTesting/Verify_R_FS"))
if("_snaps" %in% file_lists) {file_lists <- file_lists[file_lists!="_snaps"]}
testResultsRaw_list <- lapply(file_lists,function(x)testthat::test_file(paste0(path0,"/inst/SystemTesting/Verify_R_FS/",x), reporter = testthat::ListReporter))

testResultsIndividual_list <- lapply(testResultsRaw_list,function(x) x%>%
  as_tibble() %>%
  rename(Test = test) %>%
  group_by(file, context, Test) %>%
  summarise(NumTests = first(nb),
            Passed   = sum(passed),
            Failed   = sum(failed),
            Warnings = sum(warning),
            Errors   = sum(as.numeric(error)),
            Skipped  = sum(as.numeric(skipped)),
            .groups = "drop"))
testResultsIndividual<- plyr::ldply(testResultsIndividual_list)##%>%dplyr::select(-context)
testResultsNested <- testResultsIndividual %>%
  nest_by(file, context, .key = "Results") %>%
  nest_by(file, .key = "Categories")
```

```{r resetup,include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, results = "asis")
options(knitr.table.format = "pipe")
```

## Summary
```{r WriteSummary, results = "asis"}
SummariseResults(testResultsIndividual)
```

## Details
```{r WriteResults, results = "asis"}
WriteResults(testResultsNested)
## knitr::kable(testResultsIndividual)
```

```{r eval=FALSE,include=FALSE}
path0 <- here::here()
zz <- file(paste0(path0,"/inst/SystemTests/SmokeTests/TC3.3.console.txt"), open="wt")
sink(zz,  split=TRUE)
tmp <- test_file(paste0(path0,"/inst/SystemTests/SmokeTests/","ST_TestCase3.3.R"),reporter = testthat::ListReporter)
```








## Appendix: Tests Outputs



```{r results='hold',message=FALSE,warning=FALSE,fig.show='hold'}
#file_lists <- list.files(paste0("path0","/inst/SystemTests/SmokeTests"))
##file_lists <- c("ST_TestCase3.1.R","ST_TestCase3.2.R","ST_TestCase3.3.R")
file_lists <- c("ST_TestCase3.1.R","ST_TestCase3.2.R")
testResultsRaw_list <- lapply(file_lists,function(x)testthat::test_file(paste0(path0,"/inst/SystemTests/SmokeTests/",x), reporter = testthat::ListReporter))
```



```{r}

```


```{r}

```


# References

::: {#refs}
:::
