## Data Simulation

In this article we illustrate how to simulating dose-response data using specific models from the `drc`, `MCPMod`, and `DoseFinding` packages.

### 1. **Using the `drc` Package**

The `drc` package provides functions for fitting and simulating dose-response curves. Here are a few common models you can simulate:

- **Log-Logistic Model**: A common model for dose-response relationships.
- **Log-Probit Model**: A common model for dose-response relationships.
- **Weibull Model**: Another flexible model for dose-response data.
- **Weibull Model**: Another flexible model for dose-response data.

#### Example Code

```{r}
# Load necessary library
library(drc)

# Define parameters for the logistic model
params <- c(ED50 = 10, slope = 1, upper = 100, lower = 0)

# Simulate dose-response data
set.seed(123)
doses <- seq(0, 20, by = 1)
responses <- drc::rdrm(drc::LL.4(params), newdata = data.frame(dose = doses))

# Create a data frame
simulated_data <- data.frame(Dose = doses, Response = responses)
```

### 2. **Using the `MCPMod` Package**

The `MCPMod` package is designed for modeling and testing dose-response relationships using a combination of models. You can simulate data based on specific dose-response models defined in the package.

#### Example Code

```{r eval=FALSE}
# Load necessary library
library(MCPMod)

# Define a dose-response model (e.g., Emax model)
model <- EmaxModel()

# Simulate dose-response data
set.seed(123)
doses <- seq(0, 20, by = 2)
simulated_data <- simulate(model, n = 100, dose = doses)

# The simulated_data will include doses and corresponding responses
```

### 3. **Using the `DoseFinding` Package**

The `DoseFinding` package provides tools for designing dose-finding studies and includes functions for simulating dose-response data.

#### Example Code

```{r}
# Load necessary library
library(DoseFinding)

# Define a logistic dose-response function
logistic_function <- function(dose, lower = 0, upper = 100, ED50 = 10, slope = 1) {
  lower + (upper - lower) / (1 + (dose / ED50)^(-slope))
}

# Simulate dose-response data
set.seed(123)
doses <- seq(0, 20, by = 2)
responses <- logistic_function(doses) + rnorm(length(doses), mean = 0, sd = 5)  # Adding noise

# Create a data frame
simulated_data <- data.frame(Dose = doses, Response = responses)
prelimPlot2(simulated_data)
```

### 4. **Using Custom Models**

You can also simulate dose-response data using custom equations. For example, you can define your own logistic or exponential models and generate data based on those.

#### Example Code for Custom Logistic Model

```{r}
# Custom logistic function
custom_logistic <- function(dose, lower = 0, upper = 100, ED50 = 10, slope = 1) {
  lower + (upper - lower) / (1 + exp(-slope * (dose - ED50)))
}

# Simulate data
set.seed(123)
doses <- seq(0, 20, by = 1)
responses <- custom_logistic(doses) + rnorm(length(doses), mean = 0, sd = 5)  # Adding noise

# Create a data frame
simulated_data <- data.frame(Dose = doses, Response = responses)
prelimPlot2(simulated_data)
```

### Summary

You can simulate dose-response data using various models from the `drc`, `MCPMod`, and `DoseFinding` packages, or by creating custom models. The choice of model will depend on the specific characteristics of the data you wish to simulate, such as the expected shape of the dose-response curve, the presence of noise, and the desired parameters. Each of the examples provided can be customized further to suit your simulation needs.
