---
title: "A Notes on Power"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(drcHelper)
library(dplyr)
```


## Concepts 

In statistical analysis, tests can be categorized as liberal or conservative based on their propensity to reject the null hypothesis. Here are examples of both types of tests:

### Liberal Tests

1. **Least Significant Difference (LSD) Test**:
   - Used for pairwise comparisons following ANOVA.
   - Does not adjust for multiple comparisons, making it more prone to Type I errors.

2. **Unadjusted t-tests for Multiple Comparisons**:
   - Performing multiple t-tests without any correction for multiple testing increases the chance of Type I errors.

3. **Chi-Square Test Without Yates' Continuity Correction**:
   - In small sample sizes, not applying Yates' continuity correction can make the test more liberal.

### Conservative Tests

1. **Bonferroni Correction**:
   - Adjusts the significance level for multiple comparisons by dividing the alpha level by the number of tests.
   - Reduces the likelihood of Type I errors but can be overly conservative, increasing the risk of Type II errors.

2. **Tukey's Honest Significant Difference (HSD) Test**:
   - Used for multiple comparisons following ANOVA.
   - Controls the family-wise error rate, making it more conservative than LSD.

3. **Holm-Bonferroni Method**:
   - A stepwise approach to control the family-wise error rate.
   - Less conservative than Bonferroni but still provides strong control over Type I errors.

4. **Yates' Continuity Correction for Chi-Square Tests**:
   - Applied to 2x2 contingency tables to make the test more conservative, especially with small sample sizes.

### Implications of Choosing Liberal vs. Conservative Tests

- **Liberal Tests**: These are useful when the primary concern is maximizing the power to detect differences, and the cost of Type I errors is low. However, they should be used with caution when multiple comparisons are involved.

- **Conservative Tests**: These are preferred when controlling for Type I errors is crucial, such as in confirmatory studies where false positives could lead to incorrect scientific conclusions. However, they may increase the risk of Type II errors, potentially missing true effects.

Choosing between liberal and conservative tests depends on the research context, study design, and the balance between Type I and Type II error risks. Researchers should carefully consider these factors when selecting statistical methods for their analyses.


## A Powerful Test

No, a more powerful test is not necessarily the same as a more liberal test. These terms refer to different aspects of statistical testing:

### Power of a Test

- **Definition**: The power of a statistical test is the probability that the test correctly rejects a false null hypothesis (i.e., it detects an effect when there is one). Higher power means a greater ability to detect true effects.
- **Factors Influencing Power**: Larger sample sizes, larger effect sizes, and lower variability increase the power of a test. The significance level (alpha) also affects power; a higher alpha can increase power but also increases the risk of Type I errors.

### Liberal Test

- **Definition**: A liberal test is one that has a higher probability of rejecting the null hypothesis, which can lead to more Type I errors (false positives). It is less stringent in terms of controlling for these errors.
- **Characteristics**: Typically, liberal tests have a higher Type I error rate than the nominal level, making them more prone to finding statistically significant results even when there is no true effect.

### Differences Between Power and Liberalism

1. **Power**:
   - Focuses on the test's ability to detect true effects (avoiding Type II errors, or false negatives).
   - A test can be powerful without being liberal if it maintains the correct Type I error rate while maximizing the ability to detect true effects.

2. **Liberalism**:
   - Focuses on the test's tendency to reject the null hypothesis, potentially at the expense of increasing Type I errors.
   - A liberal test may appear more "powerful" in terms of rejecting the null hypothesis, but this can be misleading because it might also reject the null when it is true.

### Conclusion

While both power and liberalism relate to the outcomes of hypothesis testing, they address different types of errors and have different implications for statistical decision-making. Ideally, a test should be both powerful and correctly control the Type I error rate, achieving a balance between detecting true effects and avoiding false positives.


## Nonparametric Tests

### Many-to-one Wilcoxon v.s. Dunn

The many-to-one version of Dunn's test is more powerful compaired to multiple wilcoxon rank sum test because it maintains the relationship between all groups throughout the analysis by using ranks from the complete dataset to compute test statistics. This integrated approach provides better statistical efficiency compared to the separate pairwise comparisons in Wilcoxon tests.

1. uses complete ranking information rather than pairwise rankings.
2. more efficiently uses the data structure by considering all groups simultaneously
3. reduced multiple comparison penalty in many-to-one setup
4. Dunn's test handles tied ranks more efficiently across the entire dataset, while Wilcoxon rank sum test processes ties separately for each comparison.

This reminds me of how ANOVA is generally more powerful than multiple t-tests - it's a similar principle of using all available information simultaneously rather than breaking it into pieces.


```{r}
library(PMCMRplus)
## Data set PlantGrowth
## Global test

prelimPlot3(PlantGrowth %>% mutate(Dose=group,Response=weight))
kruskalTest(weight ~ group, data = PlantGrowth)
```
The global test is significant (p < 0.05), so post-hoc tests are justified.

```{r}
# --- 2. Run All Requested Many-to-One Comparison Tests ---

# We will suppress warnings about ties, as they are expected and handled by the functions.
# Each function returns an object where the p-values are stored.

# Conover's test with "single-step" adjustment (specific for many-to-one)
ans_conover_ss <- suppressWarnings(
  kwManyOneConoverTest(weight ~ group, data = PlantGrowth, p.adjust.method = "single-step")
)

# Conover's test with "holm" adjustment (general-purpose)
ans_conover_holm <- suppressWarnings(
  kwManyOneConoverTest(weight ~ group, data = PlantGrowth, p.adjust.method = "holm")
)

# Dunn's test with "holm" adjustment
ans_dunn <- suppressWarnings(
  kwManyOneDunnTest(weight ~ group, data = PlantGrowth, p.adjust.method = "holm")
)

# Nemenyi's test with "holm" adjustment
ans_nemenyi <- suppressWarnings(
  kwManyOneNdwTest(weight ~ group, data = PlantGrowth, p.adjust.method = "holm")
)

# Many-one U-test (based on Mann-Whitney U) with "holm" adjustment
ans_utest <- suppressWarnings(
  manyOneUTest(weight ~ group, data = PlantGrowth, p.adjust.method = "holm")
)

# Chen's test (for heteroscedastic data) with "holm" adjustment
ans_chen <- suppressWarnings(
  chenTest(weight ~ group, data = PlantGrowth, p.adjust.method = "holm")
)


# --- 3. Extract and Combine Results into a Single Table ---

# Create a list of the result objects with descriptive names
test_results <- list(
  Conover_SingleStep = ans_conover_ss,
  Conover_Holm = ans_conover_holm,
  Dunn_Holm = ans_dunn,
  Nemenyi_Holm = ans_nemenyi,
  ManyOneU_Holm = ans_utest,
  Chen_Holm = ans_chen
)

# Use `purrr::map` to iterate through the list, extracting p-values into data frames
# `tibble::rownames_to_column` is used to turn the comparison names (e.g., "trt1") into a proper column.
#p_value_list <- purrr::imap(test_results, ~ .x$p.value %>% 
#                       as.data.frame() %>%
#                      tibble::rownames_to_column("Comparison") %>%
#                       rename(!!.y := ctrl)) # Rename the p-value column to the test name

# Get the standard comparison group names from the data
# This will be c("trt1", "trt2")
comparison_levels <- setdiff(levels(PlantGrowth$group), "ctrl")

# Create a list to store the cleaned data frames
cleaned_p_value_list <- list()

# Loop through each test result to extract p-values robustly
# --- 2. THE CORRECTED EXTRACTION AND COMBINING LOGIC ---

p_value_list <- purrr::imap(test_results, ~ {
  # Start with a clean data frame with just the comparison names
  df <- data.frame(Comparison = comparison_levels)
  
  # Extract the p-value matrix and get the value column name
  p_matrix <- .x$p.value
  p_col_name <- colnames(p_matrix)[1]

  # --- This is the crucial fix ---
  # Convert the p-value matrix to a data frame and use pull() to ensure 
  # it becomes a simple vector, not a matrix-column.
  p_vector <- as.data.frame(p_matrix) %>% 
                pull(!!p_col_name)
  
  # Add the clean p-value vector to our data frame under the correct test name
  df[[.y]] <- p_vector
  
  return(df)
})

# Use `purrr::reduce` to join all the cleaned data frames by the 'Comparison' column

summary_table <- purrr::reduce(p_value_list, full_join, by = "Comparison")
summary_table
```


```{r echo=FALSE}
# --- 4. Display the Final Summary Table ---

cat("=================================================================\n")
cat("      Combined P-Values from Various Many-to-One Tests\n")
cat("=================================================================\n")
cat("Data: weight ~ group from PlantGrowth dataset\n")
cat("Control group: 'ctrl'\n\n")
# Explicitly round all numeric columns to a reasonable precision (e.g., 5 digits)
# This removes the floating-point noise.
cleaned_summary_table <- summary_table %>%
  mutate(across(where(is.numeric), ~ round(.x, 5)))
# --- The Definitive Fix ---
# Instead of rounding to a number, we format to a string with 3 decimal places.
# This creates clean character strings ("0.355", "0.216", "0.009", etc.)
formatted_summary_table <- summary_table %>%
  mutate(across(where(is.numeric), ~ sprintf("%.3f", .x)))
# Now, kable's digit formatting will work as expected
formatted_summary_table %>% 
  knitr::kable(digits = 3)
# print(summary_table)

```

- Note that `manyOneUTest` is a function that systematically performs a Wilcoxon/Mann-Whitney U test for each treatment versus control and then corrects the resulting p-values for multiple testing.


