---
title: "EFSA Criteria"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(drcHelper)
```


## Model Comparison Criteria

Apart from generic model comparison criteria, EFSA has proposed to use the following two criteria 

**Normalized Width**

|NW   |    Rating|
|:----|:---------|
|$<0.2$ | Excellent|
|$<0.5$ | Good     |
|$<1$   | Fair     |
|$<2$   | Poor     |
|$>2$ | Bad      |

**Overlapping of EC10, 20 and 50**


|Overlapping  Conditions                  |Certainty of the Protection Level |
|:----------------------------------------|:---------------------------------|
|EC$_{10}$ < EC$_{20,low}$                | High                             |
|EC$_{20,low}$ < EC$_{10}$ < EC$_{50,low}$| Medium                           |
|EC$_{10}$ > EC$_{50,low}$                | Fair                             |




## BMD Guidance 2012 Criteria

## BMD Guidance 2022 Criteria

## US EPA BMDS Criteria








| Possible Guidelines                                                                                                       |
|:--------------------------------------------------------------------------------------------------------------------------|
| OECD guideline 202 (Daphnia sp. Acute Immobilisation Test) - 2004                                                         |
| OECD guideline 202 (Daphnia sp. Acute Immobilisation Test) - 1984                                                         |
| OECD guideline 202 (Daphnia sp. Acute Immobilisation Test) - 1981                                                         |
| OCSPP 850.1010 Aquatic Invertebrate Acute Toxicity Test, Freshwater Daphnids - 2016                                       |
| OPPTS 850.1010 Aquatic Invertebrate Acute Toxicity Test, Freshwater Daphnids - 1996                                       |
| EPA OPP 72-2 (Aquatic Invertebrate Acute Toxicity Test)                                                                   |
| ISO 6341 (Water quality - Determination of the Inhibition of the Mobility of Daphnia magna Straus (Cladocera, Crustacea)) |
| EU Method C.2 (Acute Toxicity for Daphnia)                                                                                |
| Japanese MAFF notification No.12-Nousan-8147 - 2000                                                                       |
| Japanese MAFF notification No. 30-Shouan-6278 - 2019                                                                      |
