---
title: "Equivalence Testing"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(drcHelper)
```

## Introduction

Equivalence testing is a statistical testing approach used to determine whether two treatments or interventions produce effects that are practically the same within a predefined margin of difference. This method is particularly useful in fields like pharmaceuticals, where demonstrating that a new drug is not worse than an existing one by more than a specified margin is crucial.

Equivalence testing is closely related to difference testing. While the latter focuses on identifying whether a difference exists by controlling the false positive error, the former aims to show that any difference is within a predefined acceptable range, by controlling the false negative error in the difference testing scenario.

he design and interpretation of equivalence tests can be more complex than traditional difference tests, requiring careful consideration of the equivalence margin, which could be so called biologically relevant effect size.

The null and alternative hypotheses in equivalence tests are

$$
H_0: | \mu_1 - \mu_0 | > \Delta
$$
$$
H_1: | \mu_1 - \mu_0 | \leq \Delta
$$
 where $\mu_1$ and $\mu_0$ are the means of the two treatments, and $\Delta$ is the equivalence margin.
 
 There are also non-inferiority or non-superiority cases, where the nulls are $\mu_1 - \mu_0 <= - \Delta$ or $\mu_1 - \mu_0 > \Delta$
 
Sometimes the tests can be formulated on standardized differences or means. Depending on how you frame the problem, the multiplicative model for ratios-to-control comparisons could be used. Related approach is available
 
The $\alpha$ level in equivalence testing is the probability of making a Type I error, which occurs when the null hypothesis is incorrectly rejected. This means concluding that the treatments are equivalent when they are not.

The $\beta$ in difference testing is the probability of making a Type II error, which occurs when the null hypothesis is not rejected when it should be. This means failing to detect a difference (concluding equivalence) when one actually exists.

In equivalence testing, a low alpha level is crucial to ensure that the conclusion of equivalence is reliable. Conversely, in difference testing, a low beta level is important to ensure that a true difference is detected.

On the other hand, the $\alpha$ level in difference testing is playing similarly a complementary role to the $\beta$ in equivalence testing. 

## Impacts of changing from difference to equivalence testing

- Equivalence testing often requires larger sample sizes to achieve sufficient power, which can be resource-intensive.
- The design and interpretation of equivalence tests can be more complex than traditional difference tests, requiring careful consideration of the "equivalence margin".

## An example using chick weights dataset

```{r}

```

```{r}
if(require(equivUMP))
  # compare two feed from chickwts dataset
  data("chickwts")
  chickwts2 <- chickwts[chickwts$feed %in% c("linseed", "soybean"),]
  chickwts2$feed <- droplevels(chickwts2$feed)

  # similar but cannot be shown to be equivalent up to 0.5 sigma at 0.05 level^
  plot(weight ~ feed, data = chickwts2)
  equiv.test(weight ~ feed, data = chickwts2, eps = 0.5)
```


## References

- Dilba, G., Bretz, F., Guiard, V., and Hothorn, L. A. (2004). Simultaneous confidence intervals for ratios with applications to the comparison of several treatments with a control. Methods of Information in Medicine 43, 465–469.
- Djira G, Hasler M, Gerhard D, Segbehoe L, Schaarschmidt F (2025). _mratios: 
Ratios of Coefficients in the General Linear Model_. R package version 1.4.4,
  <https://CRAN.R-project.org/package=mratios>.
