---
title: "MQJT"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(drcHelper)
source("../knitr-setup.R")
```

/**
 Need re-write this article!!
 **/

## Multiquantal Jonckheere-Terpstra (MQJT) Test

The standard Jonckheere-Terpstra test compares groups using a single response variable. For ordinal endpoints such as NF developmental stage in AMA studies, the **Multiquantal JT (MQJT)** extends the standard approach by examining multiple quantiles of the stage distribution simultaneously.

### Motivation

A step-down JT test applied to replicate medians is the conventional approach for AMA developmental stage. However, median-only analysis may miss shifts that affect primarily the tails of the distribution. The MQJT addresses this by dichotomising the ordinal response at multiple cutpoints and applying a step-down JT at each "slice".

**Key assumption**: If there is a true treatment effect on the population, a shift in the *entire* distribution is expected. Even if replicate medians appear similar across concentrations, effects at other quantiles should be detected.

### Algorithm

The MQJT procedure (Green, Springer & Holbech 2018, ch. 9.3) works as follows:

1. **Dichotomise**: For each cutpoint $ from the set of observed control stages (excluding the minimum), create a binary indicator (\text{stage} \geq c)$ for each animal.

2. **Aggregate to replicates**: Compute the proportion of animals in each replicate (tank) with $\text{stage} \geq c$.

3. **Step-down JT**: Apply the step-down Jonckheere-Terpstra procedure to these proportions:
   - Start with all dose groups. If the JT test is significant ( < \alpha$), the NOEC at this slice is below all tested doses.
   - Remove the highest dose and re-test. Repeat until the test is no longer significant.
   - The NOEC for this slice is the highest dose remaining at the point where significance is lost.

4. **Overall NOEC**: The most conservative (lowest) NOEC across all slices is the final NOEC.

### Implementation in drcHelper

The mqjt_test() function implements this procedure using PMCMRplus::jonckheereTest() as the underlying JT engine.

```{r mqjt-example}
# Create example AMA-like data with a dose-dependent shift in stage distribution
set.seed(42)
ama_data <- data.frame(
  stage = c(
    sample(54:62, 30, replace = TRUE, prob = c(1, 1, 2, 3, 4, 5, 5, 4, 3)),
    sample(54:62, 30, replace = TRUE, prob = c(1, 2, 3, 4, 5, 5, 4, 3, 1)),
    sample(54:62, 30, replace = TRUE, prob = c(2, 3, 4, 5, 5, 4, 3, 1, 1)),
    sample(54:62, 30, replace = TRUE, prob = c(4, 5, 5, 4, 3, 2, 1, 1, 1))
  ),
  dose = factor(rep(c("Control", "0.1", "1.0", "10"), each = 30),
                levels = c("Control", "0.1", "1.0", "10")),
  tank = rep(rep(paste0("T", 1:3), each = 10), 4)
)

# Run MQJT with one-sided test for decreasing trend
res <- mqjt_test(
  data = ama_data,
  stage_col = "stage",
  dose_col = "dose",
  replicate_col = "tank",
  alternative = "less"
)
```

### Results

```{r mqjt-print}
res
```

The summary() method provides per-cutpoint detail:

```{r mqjt-summary}
summary(res)
```

### Interpretation

```{r mqjt-interpret}
# NOEC per slice
res$noec_per_slice
```

```{r mqjt-overall}
# Overall NOEC (most conservative across all slices)
res$overall_NOEC
```

The overall NOEC is the **minimum** NOEC across all cutpoints, ensuring the most protective result. If any quantile of the stage distribution shows a significant treatment-related shift at a given dose, that dose is considered the LOEC.

### Comparison: Standard JT vs. MQJT

| Aspect | Standard step-down JT | MQJT |
|--------|----------------------|------|
| Input | Single response per replicate (e.g. median stage) | Multiple quantiles of stage distribution |
| Sensitivity | May miss tail effects | Detects shifts across entire distribution |
| NOEC determination | Based on single step-down | Most conservative across all slices |
| Use case | General trend testing | Ordinal endpoints (AMA, LAGDA) |

### References

- Green JW, Springer TA, Holbech H (2018). *Statistical Analysis of Ecotoxicity Studies*. Wiley. Chapter 9, Sections 9.3.1--9.3.4.
- OECD Test Guideline 231: Amphibian Metamorphosis Assay.


MQJT(Multiquantal Jonckheere–Terpstra) is mentioned in OECD TG 231 for AMA life stage (integers ranging from 45 to 66 in general) analysis. It is identical to the standard JT test except that it is applied to several quantiles in addition to the median and decisions are made using the median p‐value.  

## The MQJT

The step-down Jonckheere-Terpstra test applied to replicate medians was conventionally used for statistical analysis of amphibian developmental stage. Some questions have been raised whether an analysis focused on the median developmental stage in each rep is sufficiently sensitive to detect statistically effects on developmental stage that are biologically important. Thus John and Tim proposed a modification on standard step-down Jonckheere-Terpstra test procedure which examines multiple quantiles (i.e. shifts in Q20, Q30, Q40, Q50, . . . , Q80) simultaneously. 

The assumption behind MQJT is that a shift in the entire distribution would be expected if there is a "true" effect on the population. An effect on a small portion of the more sensitive population was not considered as biologically relevant if the majority in that replicate were not affected. On the other hand, even if the replicate median were similar across all test concentrations / doses, if most other quantiles are different, showing and effect of the test item, then a significant effect should be identified by the test.  

Assuming a common distribution in all test concentrations (a standard assumption in
almost all statistical analyses), the difference between the nth quantile, Qn, and the
median, Q50, is constant across concentrations. Thus, the shifts in Q20, Q30, Q40, Q50, .
. ., Q80, all measure the effect of test concentrations on a segment of the frequency
distribution of stages.

Excerpt from John's "Supplementary Approach to Statistical Analysis of Developmental Stage in
Amphibians"  
The proposed method proceeds as follows:
1. Determine the 20th percentile of the stage distribution within each rep at each test
concentration.

2. Perform a Jonkheere-Terpstra trend test on these 20th percentiles across all
concentrations. Record the P-value for the trend test.

3. Repeat steps 1 and 2 for the 30th through 80th percentile.

4. Determine the median of the P-values that were obtained in steps 1-3. This
median P-value is the probability that a trend in distribution shifts across
concentrations is due to chance. If this median p-value is 0.05 or greater, testing
stops and the NOEC exceeds the highest tested concentration. Otherwise, proceed
to step (5).

5. Drop the data from the highest test concentration, and repeat steps 1-4, so that a
new median P-value is calculated for the reduced data set. If this median p-value
is 0.05 or greater, testing stops and the NOEC is the highest tested concentration
used in this analysis. Otherwise, proceed to step (6).

6. Drop the data from the highest concentration used in the current analysis, and
repeat steps 1-4, so that a new median P-value is calculated for the reduced data
set.

7. Continue dropping treatment groups until the new median P-value is no longer
statistically significant. The highest concentration at which this occurs is the noobserved-
effect-concentration (NOEC).


## References

- OECD TG 231, https://onlinelibrary.wiley.com/doi/abs/10.1002/9781119488798.ch9
