---
title: "NOEC, ECx, BMD, etc."
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(drcHelper)
```

## Current Statistical Practices

Current statistical practices typically involve the following steps: 

1. Design Experimental Study (according to corresponding test guideline)
    - Define endpoints (survival, growth, reproduction, etc.) 
    - Identify concentration range and replication

2. Data Acquisition \& Quality Check 
    - Randomization and controls 
    - Data integrity, outliers, missing values

3. Exploratory Data Analysis     
    - Plot response vs. dose 
    - Assess data distribution and variance homogeneity

4. NOEC Determination  
   - Apply ANOVA/multiple comparisons (e.g., Dunnett's) or sequential tests.  
   - Identify highest dose without significant difference

5. Dose–Response Modeling and ECx Calculation 
    - Fit a model (e.g., logistic, probit, Weibull) to the data
    - Evaluate goodness-of-fit, possible data transformations
    - Derive ECx values (EC10, EC20, etc.) using the fitted model
    - Estimations and predictions if needed

6. Uncertainty & Sensitivity Analysis
    - Confidence intervals for NOEC and ECx values
    - Comparison of model-based ECx estimates versus NOEC

7. Regulatory Risk Assessment
    - Use NOEC or ECx values along with safety factors to define acceptable exposure limits


### NOEC

– Definition: The highest tested concentration at which no statistically significant adverse effect is observed relative to the control.
– Regulatory Role: Historically used as the primary benchmark (e.g., for deriving acceptable daily intakes or environmental quality standards).

**Strengths:**

– Straightforward concept and quick to interpret
– Familiar to risk assessors and historically embedded in many regulatory frameworks

**Weaknesses:**

– Strongly dependent on the dose spacing and study design
– Does not provide information on the magnitude of effects or their progression along a dose–response curve
– Binary interpretation ("effect" vs. "no effect") can oversimplify data
– Potentially influenced by low statistical power, especially in studies with small sample sizes

### ECx and BMD

– Definition: The concentration at which a predefined percentage (x) of effect (e.g., 10\%, 20\% reduction in a biological endpoint) is observed relative to the control.

– Regulatory Role: Increasingly used to quantitatively describe the magnitude of adverse effects along the dose–response curve and considered more informative in deriving species sensitivity distributions.

**Strengths:**

– Utilizes the full data set by modeling the dose–response relationship
– Provides a more nuanced and continuous estimate along the effect continuum (e.g., EC10 is often seen as protective)
– Facilitates interpolation and extrapolation within the tested range
– Offers the possibility to quantify uncertainty via confidence intervals

**Weaknesses:**

– Heavily dependent on the choice and fit of the dose–response model
– Requires more sophisticated statistical methods and expertise
– May be sensitive to violations of model assumptions if data quality or distribution is suboptimal
– In some cases, different models may yield different ECx estimates, raising issues of model selection and validation

## Methodological Improvements

SETAC’s journals (Environmental Toxicology and Chemistry) regularly publish papers on methodological improvements in statistical evaluation of ecotoxicological studies.

- We advocate for transparent reporting of uncertainties and assumptions in model selection or model averaging.
- Integration of Bayesian methods and bootstrap techniques to better quantify uncertainty is emerging in research discussions. 
- Model averaging.



## References

- U.S. Environmental Protection Agency (EPA). EPA Guidelines for Ecological Risk Assessment – available on the EPA website. URL: https://www.epa.gov/risk

- EFSA (2013) Guidance on the environmental risk assessment of plant protection products.

- OECD Guidelines for the Testing of Chemicals detail several standardized tests for ecotoxicity and discuss both NOEC and model-based endpoints.
URL: https://www.oecd-ilibrary.org/environment/oecd-guidelines-for-the-testing-of-chemicals

