---
title: "Study Types and Templates"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tidyverse)
```

```{r setup}
library(drcHelper)
```

Statistical methods are used to assess the effects of PPPs on

- **Non-Target Plants**: Endpoints may include emergence, growth, and reproduction.
- **Aquatic Organisms**: such as fish, invertebrates, and algae.
- **Terrestrial Invertebrates**: such as earthworms, bees, and predatory insects.
- **Birds and Mammals**: Endpoints may include survival, reproduction, and behavior.

There are a few analysis templates available for certain type of studies in the first release of the accompanied shiny app for V-COP platform.

- EFX-Aquatics: Myriophyllum test (OECD 239)
- EFX-NTA: Aphidius extended lab test (IOBC)
- EFX-NTP: Seedling emergence test (OECD 208) 

For Myriophyllum test, the interested endpoints are growth rate and yield for total shoot lenght, fresh weight, and dry weight at day 14. 


For Aphidius extended lab test, the endpoints are mortality at 48h and reproductivity at 14 d, the response variable is the number of individuals or juveniles.

For Seedling emergence test, the endpoints are Seedling emergence, Survival of emerged seedlings, Plant height, and Shoot dry weight at 21 day.


```{r include=FALSE}
NTA <- read.table(text='Study type;Test organism;Design;Time;Endpoint
Ab_ext;Aleochara bilineata;Extended;not determined;Reproduction
Ar_ext;Aphidius rhopalosiphi;Extended;48 h;Mortality
;;;(14 d is the default, other value is possible);Reproduction
;;;not determined;Repellency
Ar_lab;Aphidius rhopalosiphi;Laboratory;48 h;Mortality
;;;(14 d is the default, other value is possible);Reproduction
Ar_aged;Aphidius rhopalosiphi;Aged;48 h;Mortality
;;;(14 d is the default, other value is possible);Reproduction
;;;not determined;Repellency
Cc_ext;Chrysoperla carnea;Extended;not determined;Mortality
;;;not determined;Reproduction
Cs_ext;Coccinella septempunctata;Extended;not relevant;Mortality
;;;not determined;Reproduction
Ol_ext;Orius laevigatus;Extended;7 d;Mortality
;;;not determined;Fecundity
;;;not determined;Fertility
Seeds_LeafPower;n.d.;Seeds LeafPowder;(7 d is the default, other value is possible);Mortality
Special_Studies;n.d.;Special Studies;(10 d is the default, other value is possible);Mortality
Tp_ext;Typhlodromus pyri;Extended;(7 d is the default, other value is possible);Mortality
;;;(7,9,12,14  d are the defaults, other values are possible);Reproduction
Tp_lab;Typhlodromus pyri;Laboratory;(7 d is the default, other value is possible);Mortality
;;;(7,9,12,14  d are the defaults, other values are possible);Reproduction
Tp_aged;Typhlodromus pyri;Aged;(7 d is the default, other value is possible);Mortality
;;;(7,9,12,14  d are the defaults, other values are possible);Reproduction
Tp_Munger;Typhlodromus pyri;Munger;(7 d is the default, other value is possible);Mortality
;;;(7,9,12,14  d are the defaults, other values are possible);Reproduction',header=TRUE,sep=";")

```

## Other NTA Studies

```{r eval=FALSE,include=FALSE}
NTA <- NTA %>% fill(.)
gt::gt(NTA)
```

<!--html_preserve--><div id="rxeemggklk" style="padding-left:0px;padding-right:0px;padding-top:10px;padding-bottom:10px;overflow-x:auto;overflow-y:auto;width:auto;height:auto;">
<style>#rxeemggklk table {
  font-family: system-ui, 'Segoe UI', Roboto, Helvetica, Arial, sans-serif, 'Apple Color Emoji', 'Segoe UI Emoji', 'Segoe UI Symbol', 'Noto Color Emoji';
  -webkit-font-smoothing: antialiased;
  -moz-osx-font-smoothing: grayscale;
}

#rxeemggklk thead, #rxeemggklk tbody, #rxeemggklk tfoot, #rxeemggklk tr, #rxeemggklk td, #rxeemggklk th {
  border-style: none;
}

#rxeemggklk p {
  margin: 0;
  padding: 0;
}

#rxeemggklk .gt_table {
  display: table;
  border-collapse: collapse;
  line-height: normal;
  margin-left: auto;
  margin-right: auto;
  color: #333333;
  font-size: 16px;
  font-weight: normal;
  font-style: normal;
  background-color: #FFFFFF;
  width: auto;
  border-top-style: solid;
  border-top-width: 2px;
  border-top-color: #A8A8A8;
  border-right-style: none;
  border-right-width: 2px;
  border-right-color: #D3D3D3;
  border-bottom-style: solid;
  border-bottom-width: 2px;
  border-bottom-color: #A8A8A8;
  border-left-style: none;
  border-left-width: 2px;
  border-left-color: #D3D3D3;
}

#rxeemggklk .gt_caption {
  padding-top: 4px;
  padding-bottom: 4px;
}

#rxeemggklk .gt_title {
  color: #333333;
  font-size: 125%;
  font-weight: initial;
  padding-top: 4px;
  padding-bottom: 4px;
  padding-left: 5px;
  padding-right: 5px;
  border-bottom-color: #FFFFFF;
  border-bottom-width: 0;
}

#rxeemggklk .gt_subtitle {
  color: #333333;
  font-size: 85%;
  font-weight: initial;
  padding-top: 3px;
  padding-bottom: 5px;
  padding-left: 5px;
  padding-right: 5px;
  border-top-color: #FFFFFF;
  border-top-width: 0;
}

#rxeemggklk .gt_heading {
  background-color: #FFFFFF;
  text-align: center;
  border-bottom-color: #FFFFFF;
  border-left-style: none;
  border-left-width: 1px;
  border-left-color: #D3D3D3;
  border-right-style: none;
  border-right-width: 1px;
  border-right-color: #D3D3D3;
}

#rxeemggklk .gt_bottom_border {
  border-bottom-style: solid;
  border-bottom-width: 2px;
  border-bottom-color: #D3D3D3;
}

#rxeemggklk .gt_col_headings {
  border-top-style: solid;
  border-top-width: 2px;
  border-top-color: #D3D3D3;
  border-bottom-style: solid;
  border-bottom-width: 2px;
  border-bottom-color: #D3D3D3;
  border-left-style: none;
  border-left-width: 1px;
  border-left-color: #D3D3D3;
  border-right-style: none;
  border-right-width: 1px;
  border-right-color: #D3D3D3;
}

#rxeemggklk .gt_col_heading {
  color: #333333;
  background-color: #FFFFFF;
  font-size: 100%;
  font-weight: normal;
  text-transform: inherit;
  border-left-style: none;
  border-left-width: 1px;
  border-left-color: #D3D3D3;
  border-right-style: none;
  border-right-width: 1px;
  border-right-color: #D3D3D3;
  vertical-align: bottom;
  padding-top: 5px;
  padding-bottom: 6px;
  padding-left: 5px;
  padding-right: 5px;
  overflow-x: hidden;
}

#rxeemggklk .gt_column_spanner_outer {
  color: #333333;
  background-color: #FFFFFF;
  font-size: 100%;
  font-weight: normal;
  text-transform: inherit;
  padding-top: 0;
  padding-bottom: 0;
  padding-left: 4px;
  padding-right: 4px;
}

#rxeemggklk .gt_column_spanner_outer:first-child {
  padding-left: 0;
}

#rxeemggklk .gt_column_spanner_outer:last-child {
  padding-right: 0;
}

#rxeemggklk .gt_column_spanner {
  border-bottom-style: solid;
  border-bottom-width: 2px;
  border-bottom-color: #D3D3D3;
  vertical-align: bottom;
  padding-top: 5px;
  padding-bottom: 5px;
  overflow-x: hidden;
  display: inline-block;
  width: 100%;
}

#rxeemggklk .gt_spanner_row {
  border-bottom-style: hidden;
}

#rxeemggklk .gt_group_heading {
  padding-top: 8px;
  padding-bottom: 8px;
  padding-left: 5px;
  padding-right: 5px;
  color: #333333;
  background-color: #FFFFFF;
  font-size: 100%;
  font-weight: initial;
  text-transform: inherit;
  border-top-style: solid;
  border-top-width: 2px;
  border-top-color: #D3D3D3;
  border-bottom-style: solid;
  border-bottom-width: 2px;
  border-bottom-color: #D3D3D3;
  border-left-style: none;
  border-left-width: 1px;
  border-left-color: #D3D3D3;
  border-right-style: none;
  border-right-width: 1px;
  border-right-color: #D3D3D3;
  vertical-align: middle;
  text-align: left;
}

#rxeemggklk .gt_empty_group_heading {
  padding: 0.5px;
  color: #333333;
  background-color: #FFFFFF;
  font-size: 100%;
  font-weight: initial;
  border-top-style: solid;
  border-top-width: 2px;
  border-top-color: #D3D3D3;
  border-bottom-style: solid;
  border-bottom-width: 2px;
  border-bottom-color: #D3D3D3;
  vertical-align: middle;
}

#rxeemggklk .gt_from_md > :first-child {
  margin-top: 0;
}

#rxeemggklk .gt_from_md > :last-child {
  margin-bottom: 0;
}

#rxeemggklk .gt_row {
  padding-top: 8px;
  padding-bottom: 8px;
  padding-left: 5px;
  padding-right: 5px;
  margin: 10px;
  border-top-style: solid;
  border-top-width: 1px;
  border-top-color: #D3D3D3;
  border-left-style: none;
  border-left-width: 1px;
  border-left-color: #D3D3D3;
  border-right-style: none;
  border-right-width: 1px;
  border-right-color: #D3D3D3;
  vertical-align: middle;
  overflow-x: hidden;
}

#rxeemggklk .gt_stub {
  color: #333333;
  background-color: #FFFFFF;
  font-size: 100%;
  font-weight: initial;
  text-transform: inherit;
  border-right-style: solid;
  border-right-width: 2px;
  border-right-color: #D3D3D3;
  padding-left: 5px;
  padding-right: 5px;
}

#rxeemggklk .gt_stub_row_group {
  color: #333333;
  background-color: #FFFFFF;
  font-size: 100%;
  font-weight: initial;
  text-transform: inherit;
  border-right-style: solid;
  border-right-width: 2px;
  border-right-color: #D3D3D3;
  padding-left: 5px;
  padding-right: 5px;
  vertical-align: top;
}

#rxeemggklk .gt_row_group_first td {
  border-top-width: 2px;
}

#rxeemggklk .gt_row_group_first th {
  border-top-width: 2px;
}

#rxeemggklk .gt_summary_row {
  color: #333333;
  background-color: #FFFFFF;
  text-transform: inherit;
  padding-top: 8px;
  padding-bottom: 8px;
  padding-left: 5px;
  padding-right: 5px;
}

#rxeemggklk .gt_first_summary_row {
  border-top-style: solid;
  border-top-color: #D3D3D3;
}

#rxeemggklk .gt_first_summary_row.thick {
  border-top-width: 2px;
}

#rxeemggklk .gt_last_summary_row {
  padding-top: 8px;
  padding-bottom: 8px;
  padding-left: 5px;
  padding-right: 5px;
  border-bottom-style: solid;
  border-bottom-width: 2px;
  border-bottom-color: #D3D3D3;
}

#rxeemggklk .gt_grand_summary_row {
  color: #333333;
  background-color: #FFFFFF;
  text-transform: inherit;
  padding-top: 8px;
  padding-bottom: 8px;
  padding-left: 5px;
  padding-right: 5px;
}

#rxeemggklk .gt_first_grand_summary_row {
  padding-top: 8px;
  padding-bottom: 8px;
  padding-left: 5px;
  padding-right: 5px;
  border-top-style: double;
  border-top-width: 6px;
  border-top-color: #D3D3D3;
}

#rxeemggklk .gt_last_grand_summary_row_top {
  padding-top: 8px;
  padding-bottom: 8px;
  padding-left: 5px;
  padding-right: 5px;
  border-bottom-style: double;
  border-bottom-width: 6px;
  border-bottom-color: #D3D3D3;
}

#rxeemggklk .gt_striped {
  background-color: rgba(128, 128, 128, 0.05);
}

#rxeemggklk .gt_table_body {
  border-top-style: solid;
  border-top-width: 2px;
  border-top-color: #D3D3D3;
  border-bottom-style: solid;
  border-bottom-width: 2px;
  border-bottom-color: #D3D3D3;
}

#rxeemggklk .gt_footnotes {
  color: #333333;
  background-color: #FFFFFF;
  border-bottom-style: none;
  border-bottom-width: 2px;
  border-bottom-color: #D3D3D3;
  border-left-style: none;
  border-left-width: 2px;
  border-left-color: #D3D3D3;
  border-right-style: none;
  border-right-width: 2px;
  border-right-color: #D3D3D3;
}

#rxeemggklk .gt_footnote {
  margin: 0px;
  font-size: 90%;
  padding-top: 4px;
  padding-bottom: 4px;
  padding-left: 5px;
  padding-right: 5px;
}

#rxeemggklk .gt_sourcenotes {
  color: #333333;
  background-color: #FFFFFF;
  border-bottom-style: none;
  border-bottom-width: 2px;
  border-bottom-color: #D3D3D3;
  border-left-style: none;
  border-left-width: 2px;
  border-left-color: #D3D3D3;
  border-right-style: none;
  border-right-width: 2px;
  border-right-color: #D3D3D3;
}

#rxeemggklk .gt_sourcenote {
  font-size: 90%;
  padding-top: 4px;
  padding-bottom: 4px;
  padding-left: 5px;
  padding-right: 5px;
}

#rxeemggklk .gt_left {
  text-align: left;
}

#rxeemggklk .gt_center {
  text-align: center;
}

#rxeemggklk .gt_right {
  text-align: right;
  font-variant-numeric: tabular-nums;
}

#rxeemggklk .gt_font_normal {
  font-weight: normal;
}

#rxeemggklk .gt_font_bold {
  font-weight: bold;
}

#rxeemggklk .gt_font_italic {
  font-style: italic;
}

#rxeemggklk .gt_super {
  font-size: 65%;
}

#rxeemggklk .gt_footnote_marks {
  font-size: 75%;
  vertical-align: 0.4em;
  position: initial;
}

#rxeemggklk .gt_asterisk {
  font-size: 100%;
  vertical-align: 0;
}

#rxeemggklk .gt_indent_1 {
  text-indent: 5px;
}

#rxeemggklk .gt_indent_2 {
  text-indent: 10px;
}

#rxeemggklk .gt_indent_3 {
  text-indent: 15px;
}

#rxeemggklk .gt_indent_4 {
  text-indent: 20px;
}

#rxeemggklk .gt_indent_5 {
  text-indent: 25px;
}
</style>
<table class="gt_table" data-quarto-disable-processing="false" data-quarto-bootstrap="false">
  <thead>
    
    <tr class="gt_col_headings">
      <th class="gt_col_heading gt_columns_bottom_border gt_left" rowspan="1" colspan="1" scope="col" id="Study.type">Study.type</th>
      <th class="gt_col_heading gt_columns_bottom_border gt_left" rowspan="1" colspan="1" scope="col" id="Test.organism">Test.organism</th>
      <th class="gt_col_heading gt_columns_bottom_border gt_left" rowspan="1" colspan="1" scope="col" id="Design">Design</th>
      <th class="gt_col_heading gt_columns_bottom_border gt_left" rowspan="1" colspan="1" scope="col" id="Time">Time</th>
      <th class="gt_col_heading gt_columns_bottom_border gt_left" rowspan="1" colspan="1" scope="col" id="Endpoint">Endpoint</th>
    </tr>
  </thead>
  <tbody class="gt_table_body">
    <tr><td headers="Study.type" class="gt_row gt_left">Ab_ext</td>
<td headers="Test.organism" class="gt_row gt_left">Aleochara bilineata</td>
<td headers="Design" class="gt_row gt_left">Extended</td>
<td headers="Time" class="gt_row gt_left">not determined</td>
<td headers="Endpoint" class="gt_row gt_left">Reproduction</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left">Ar_ext</td>
<td headers="Test.organism" class="gt_row gt_left">Aphidius rhopalosiphi</td>
<td headers="Design" class="gt_row gt_left">Extended</td>
<td headers="Time" class="gt_row gt_left">48 h</td>
<td headers="Endpoint" class="gt_row gt_left">Mortality</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left"></td>
<td headers="Test.organism" class="gt_row gt_left"></td>
<td headers="Design" class="gt_row gt_left"></td>
<td headers="Time" class="gt_row gt_left">(14 d is the default, other value is possible)</td>
<td headers="Endpoint" class="gt_row gt_left">Reproduction</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left"></td>
<td headers="Test.organism" class="gt_row gt_left"></td>
<td headers="Design" class="gt_row gt_left"></td>
<td headers="Time" class="gt_row gt_left">not determined</td>
<td headers="Endpoint" class="gt_row gt_left">Repellency</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left">Ar_lab</td>
<td headers="Test.organism" class="gt_row gt_left">Aphidius rhopalosiphi</td>
<td headers="Design" class="gt_row gt_left">Laboratory</td>
<td headers="Time" class="gt_row gt_left">48 h</td>
<td headers="Endpoint" class="gt_row gt_left">Mortality</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left"></td>
<td headers="Test.organism" class="gt_row gt_left"></td>
<td headers="Design" class="gt_row gt_left"></td>
<td headers="Time" class="gt_row gt_left">(14 d is the default, other value is possible)</td>
<td headers="Endpoint" class="gt_row gt_left">Reproduction</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left">Ar_aged</td>
<td headers="Test.organism" class="gt_row gt_left">Aphidius rhopalosiphi</td>
<td headers="Design" class="gt_row gt_left">Aged</td>
<td headers="Time" class="gt_row gt_left">48 h</td>
<td headers="Endpoint" class="gt_row gt_left">Mortality</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left"></td>
<td headers="Test.organism" class="gt_row gt_left"></td>
<td headers="Design" class="gt_row gt_left"></td>
<td headers="Time" class="gt_row gt_left">(14 d is the default, other value is possible)</td>
<td headers="Endpoint" class="gt_row gt_left">Reproduction</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left"></td>
<td headers="Test.organism" class="gt_row gt_left"></td>
<td headers="Design" class="gt_row gt_left"></td>
<td headers="Time" class="gt_row gt_left">not determined</td>
<td headers="Endpoint" class="gt_row gt_left">Repellency</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left">Cc_ext</td>
<td headers="Test.organism" class="gt_row gt_left">Chrysoperla carnea</td>
<td headers="Design" class="gt_row gt_left">Extended</td>
<td headers="Time" class="gt_row gt_left">not determined</td>
<td headers="Endpoint" class="gt_row gt_left">Mortality</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left"></td>
<td headers="Test.organism" class="gt_row gt_left"></td>
<td headers="Design" class="gt_row gt_left"></td>
<td headers="Time" class="gt_row gt_left">not determined</td>
<td headers="Endpoint" class="gt_row gt_left">Reproduction</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left">Cs_ext</td>
<td headers="Test.organism" class="gt_row gt_left">Coccinella septempunctata</td>
<td headers="Design" class="gt_row gt_left">Extended</td>
<td headers="Time" class="gt_row gt_left">not relevant</td>
<td headers="Endpoint" class="gt_row gt_left">Mortality</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left"></td>
<td headers="Test.organism" class="gt_row gt_left"></td>
<td headers="Design" class="gt_row gt_left"></td>
<td headers="Time" class="gt_row gt_left">not determined</td>
<td headers="Endpoint" class="gt_row gt_left">Reproduction</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left">Ol_ext</td>
<td headers="Test.organism" class="gt_row gt_left">Orius laevigatus</td>
<td headers="Design" class="gt_row gt_left">Extended</td>
<td headers="Time" class="gt_row gt_left">7 d</td>
<td headers="Endpoint" class="gt_row gt_left">Mortality</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left"></td>
<td headers="Test.organism" class="gt_row gt_left"></td>
<td headers="Design" class="gt_row gt_left"></td>
<td headers="Time" class="gt_row gt_left">not determined</td>
<td headers="Endpoint" class="gt_row gt_left">Fecundity</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left"></td>
<td headers="Test.organism" class="gt_row gt_left"></td>
<td headers="Design" class="gt_row gt_left"></td>
<td headers="Time" class="gt_row gt_left">not determined</td>
<td headers="Endpoint" class="gt_row gt_left">Fertility</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left">Seeds_LeafPower</td>
<td headers="Test.organism" class="gt_row gt_left">n.d.</td>
<td headers="Design" class="gt_row gt_left">Seeds LeafPowder</td>
<td headers="Time" class="gt_row gt_left">(7 d is the default, other value is possible)</td>
<td headers="Endpoint" class="gt_row gt_left">Mortality</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left">Special_Studies</td>
<td headers="Test.organism" class="gt_row gt_left">n.d.</td>
<td headers="Design" class="gt_row gt_left">Special Studies</td>
<td headers="Time" class="gt_row gt_left">(10 d is the default, other value is possible)</td>
<td headers="Endpoint" class="gt_row gt_left">Mortality</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left">Tp_ext</td>
<td headers="Test.organism" class="gt_row gt_left">Typhlodromus pyri</td>
<td headers="Design" class="gt_row gt_left">Extended</td>
<td headers="Time" class="gt_row gt_left">(7 d is the default, other value is possible)</td>
<td headers="Endpoint" class="gt_row gt_left">Mortality</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left"></td>
<td headers="Test.organism" class="gt_row gt_left"></td>
<td headers="Design" class="gt_row gt_left"></td>
<td headers="Time" class="gt_row gt_left">(7,9,12,14  d are the defaults, other values are possible)</td>
<td headers="Endpoint" class="gt_row gt_left">Reproduction</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left">Tp_lab</td>
<td headers="Test.organism" class="gt_row gt_left">Typhlodromus pyri</td>
<td headers="Design" class="gt_row gt_left">Laboratory</td>
<td headers="Time" class="gt_row gt_left">(7 d is the default, other value is possible)</td>
<td headers="Endpoint" class="gt_row gt_left">Mortality</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left"></td>
<td headers="Test.organism" class="gt_row gt_left"></td>
<td headers="Design" class="gt_row gt_left"></td>
<td headers="Time" class="gt_row gt_left">(7,9,12,14  d are the defaults, other values are possible)</td>
<td headers="Endpoint" class="gt_row gt_left">Reproduction</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left">Tp_aged</td>
<td headers="Test.organism" class="gt_row gt_left">Typhlodromus pyri</td>
<td headers="Design" class="gt_row gt_left">Aged</td>
<td headers="Time" class="gt_row gt_left">(7 d is the default, other value is possible)</td>
<td headers="Endpoint" class="gt_row gt_left">Mortality</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left"></td>
<td headers="Test.organism" class="gt_row gt_left"></td>
<td headers="Design" class="gt_row gt_left"></td>
<td headers="Time" class="gt_row gt_left">(7,9,12,14  d are the defaults, other values are possible)</td>
<td headers="Endpoint" class="gt_row gt_left">Reproduction</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left">Tp_Munger</td>
<td headers="Test.organism" class="gt_row gt_left">Typhlodromus pyri</td>
<td headers="Design" class="gt_row gt_left">Munger</td>
<td headers="Time" class="gt_row gt_left">(7 d is the default, other value is possible)</td>
<td headers="Endpoint" class="gt_row gt_left">Mortality</td></tr>
    <tr><td headers="Study.type" class="gt_row gt_left"></td>
<td headers="Test.organism" class="gt_row gt_left"></td>
<td headers="Design" class="gt_row gt_left"></td>
<td headers="Time" class="gt_row gt_left">(7,9,12,14  d are the defaults, other values are possible)</td>
<td headers="Endpoint" class="gt_row gt_left">Reproduction</td></tr>
  </tbody>
  
  
</table>
</div><!--/html_preserve-->
