---
title: "Trend Testing"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(drcHelper)
source("../knitr-setup.R")
```

## Background

**What is a trend test**

A trend test is used to detect a monotonic trend across ordered groups. It's particularly useful in dose-response studies to identify whether increasing doses lead to a systematic change in response.

The prerequisite for a trend test is that the relationship is monotonic, either consistently increasing or decreasing. These trend tests are specifically designed to detect trends, making it a bit more powerful than general tests when a trend exists.

The disadvantage is that it does not consider the magnitude of the change across the ordered groups, leading to unexpected identification of significance when there It is also less informative or even give counter-intuitive results if the relationship is not monotonic or if the groups are not naturally ordered.


## Some concerns

In lots of the decision flowcharts, step down trend test are performed after a monotonicity test passes, which is logically strange, since monotonicity test itself is also a type of trend test. For example, jonckeehre terpstra test after linear contrast is significant in monotonicity test or quadratic contrast is significant in monotonicity test. 


## Trend Test by Testing Linear and Quadratic Contrasts

Generate mock data for testing

```{r}
library(ggplot2)
mock_data <- data.frame(
  treatment_var = factor(rep(c("Control", "Dose1", "Dose2", "Dose3"), each = 10)),
  response_var = c(rnorm(10, mean = 5), rnorm(10, mean = 7), rnorm(10, mean = 8), rnorm(10, mean = 10))
)
ggplot(mock_data,aes(x=treatment_var,y=response_var))+geom_point()
```

```{r}
result <- monotonicityTest(mock_data, "treatment_var", "response_var")
result
```


## Step-Down Trend Test

A step-down trend test is a sequential testing procedure used to identify the lowest dose level at which a significant trend begins. The process works as follows:

1. Start with all dose groups and test for a trend.
2. If significant, remove the highest dose group and test again.
3. Continue this process until either:
   - The test is no longer significant, indicating you've found the lowest effective dose
   - Only two groups remain (typically control and lowest dose)

This approach helps identify the No Observed Effect Concentration (NOEC) or Lowest Observed Effect Concentration (LOEC) in dose-response studies.

The function `PMCMRplus::stepDownTrendTest` PMCMRplus implements this strategy for continuous data using various trend tests. However, the stepDownTrendTest function does not follow the NOEC testing paradigm of starting with a full model and removing the highest dose. Instead, it builds the model upwards from the control. For binomial survival data, we adapt this approach using the Cochran-Armitage test, which is specifically designed for binary outcomes across ordered groups, the function for performing step-down CA is called `drcHelper::stepDownTrendTestBinom`. There is also `drcHelper::step_down_RSCABS` and `drcHelper::stepDownRSCABS` for ordinal data. Please refere to the relevant data page for usage demo and explanations. 


```{r echo=FALSE}
data(jdata)
jdata$X <- factor(jdata$X)

# --- 1. The Official PMCMRplus::stepDownTrendTest ---
# This is our reference.
official_result <- PMCMRplus::stepDownTrendTest(Y ~ X, data = jdata, test = "jonckheereTest")
cat("--- PMCMRplus::stepDownTrendTest Output ---\n")
print(official_result)
```
```{r}
# --- 2. Manually Recreating the PMCMRplus "Step-Up" Logic ---
# This manually implements the logic inside the function to prove our understanding.
cat("\n--- Manual Recreation of PMCMRplus 'Step-Up' Logic ---\n")
levels_X <- levels(jdata$X)
# Loop from i=2 to 4
for (i in 2:length(levels_X)) {
  # Create a subset of data with groups from 1 to i
  subset_data <- subset(jdata, X %in% levels_X[1:i])
  
  # Run the test on this subset
  res <- PMCMRplus::jonckheereTest(Y ~ X, data = subset_data)
  
  cat(sprintf("Test on Groups %s vs 1: p-value = %.6f\n", i, res$p.value))
}


# --- 3. The Ecotoxicology "Step-Down" Logic (Your Manual Approach) ---
# This starts with all doses and removes the highest.
cat("\n--- Manual Recreation of Ecotoxicology 'Step-Down' Logic ---\n")
# Loop from i=4 down to 2
for (i in length(levels_X):2) {
  # Create a subset of data with groups from 1 to i
  subset_data <- subset(jdata, X %in% levels_X[1:i])
  
  # Run the test on this subset
  res <- PMCMRplus::jonckheereTest(Y ~ X, data = subset_data)
  
  cat(sprintf("Test on Groups 1 to %s: p-value = %.6f\n", i, res$p.value))
}
```
Therefore, in the `drcHelper`, a wrapper function is implemented to give the results in correct meaningful order. 


```{r}

# --- Run our NEW WRAPPER function ---
cat("\n=========================================\n")
cat("   Output from our NEW WRAPPER function\n")
cat("=========================================\n")
wrapper_output <- stepDownTrendTest_NOEC(Y ~ X, data = jdata, test = "jonckheereTest")
print(wrapper_output)
```

```{r}
noec_output <- noec_from_trend_test(Response ~ Dose, data = dat_medium, test = "jonckheereTest")
print(noec_output)
```


# Mon

In regulatory ecotox and pharma settings, the logic “test monotonicity first, then apply a trend test to derive a NOEC/MED” comes from how these procedures are constructed and what they are trying to protect you from statistically. 

**References**

- Amaratunga, D., & Ge, N. (1998). Step-down trend tests for identifying the minimum effective dose. Journal of Biopharmaceutical Statistics, 8(1), 151–162. https://doi.org/10.1080/10543409808835228
- https://www.tandfonline.com/doi/abs/10.1080/10543409808835228
- https://rdrr.io/cran/PMCMRplus/man/tamhaneDunnettTest.html

---

## 1. Context: NOEC/MED from multiple doses

**The problem:**  
You have several dose groups and a control, and you want to determine the **NOEC** (No Observed Effect Concentration) or **MED** (Minimum Effective Dose). This is typically done via:

- Multiple comparisons vs. control (e.g., Dunnett-type tests), or  
- Step-down procedures that incorporate a **trend test** across doses.

Tamhane, Hochberg, and Dunnett (and follow‑up work) studied procedures that:

1. Order doses from high to low.
2. Test for an effect starting at the highest dose.
3. Step down until you reach the first dose that is not significantly different from control → that dose is the NOEC (or one step above is the MED, depending on convention).

Amaratungs and Ge (2007) discussed some variants of this idea:

- **SD2L**: step-down procedure using a **simple linear contrast–based trend test** at each step.  
- **SD2B**: same step-down structure, but replaces the linear trend test with **Bartholomew’s test** (a more powerful ordered-mean test).  
- **SD2W**: replaces the trend test with **Welch’s test** to handle heteroscedasticity.

They show that SD2B (Bartholomew) and SD2W (Welch) often outperform SD2L, sometimes by >10% in efficiency.

---

## 2. Why a monotonicity / trend test at all?

The key assumption behind these step-down NOEC/MED procedures is that the **true dose–response is monotone** (typically non-decreasing with dose). Under monotonicity:

- Higher doses should not have *smaller* mean responses than lower doses (for a harmful effect).
- A “trend test” (linear contrast, Bartholomew, Williams, etc.) is specifically designed to detect such **ordered** alternatives more efficiently than general ANOVA-type tests.

If the response is monotone, a trend test:

- Has higher power to detect a dose–response relationship.
- Produces a more interpretable pattern for defining a minimum effective dose.

If the response is **not** monotone (e.g., U‑shaped, hormesis, random fluctuations), then:

- A monotone trend test is mis-specified and may be misleading.
- The step-down logic (“once you lose significance, all lower doses are assumed non-effective”) can fail.

Hence, one important methodological question is:  
**Should we first check whether the data are consistent with a monotone pattern before applying a monotone trend-based step-down procedure to derive NOEC/MED?**

---

## 3. The role of a monotonicity test before a trend-based NOEC procedure

Performing a monotonicity test before the trend test serves several purposes:

### 3.1. Model check: Is a monotone dose–response plausible?

Trend-based step-down procedures (SD2L, SD2B, SD2W, Williams-type tests, etc.) assume that:

- The expected means are ordered by dose (e.g., μ₀ ≤ μ₁ ≤ μ₂ ≤ … for an increasing effect).

A **monotonicity test** (e.g., Bartholomew’s test, Williams’ test, or an order-restricted test) is effectively a test of:

- H₀: μ₀ = μ₁ = … = μ_k  
  vs.  
- H₁: μ₀ ≤ μ₁ ≤ … ≤ μ_k with at least one strict inequality.

If monotonicity is rejected, then:

- The data do not support the assumption of a simple monotone dose–response.
- Using a monotone trend-based step-down NOEC procedure becomes questionable.
- You might instead:
  - Use general multiple comparisons vs. control (e.g., Dunnett, Tamhane–Dunnett, or heteroscedastic variants), or  
  - Move to model-based approaches (e.g., parametric dose–response models, non-monotone models).

### 3.2. Control of Type I error under the intended model

The Tamhane et al. paper and the follow-up (Amaratungs and Ge 2007) evaluate procedures under **balanced normal homoscedastic** (and later unbalanced/heteroscedastic) designs, assuming
