---
title: "Verification and Validation of NOEC Derivation by MCP"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Multiple Comparison Tests



```{r setup}
library(drcHelper)
library(PMCMRplus)
```


## Dunnett's Test

`glht` pulls its P-values out of a multivariate t distribution. mvtnorm::pmvt is called with the observed t statistics and correlation matrix (actual code here). DescTools::DunnettTest does roughly the same. emmeans uses a close approximation to the Dunnett adjustment. 

```{r}

```




## Williams' Test

PMCMRplus::williamsTest produces more accurate results since it is based on the isotonic regression.

Both functions use tabulated critical values provided by Williams (only for a=0.05). 

The `williamsTest` function from **PMCMRplus** produces the same results as in `williamsTest_JG`

```{r}
x <- c(106, 114, 116, 127, 145,110, 125, 143, 148, 151,
136, 139, 149, 160, 174)
g <- gl(3,5)
levels(g) <- c("0", "I", "II")
PMCMRplus::williamsTest(x ~ g)
williamsTest_JG(data.frame(treatment_var = g,response_var=x),"response_var","treatment_var",direction="increasing")

```


## Tips

Note that to install "PMCMRplus" properly under ubuntu (Platform: x86_64-pc-linux-gnu), you need probably install libmpfr-dev first. 

```
sudo apt-get install libmpfr-dev
```


