---
title: "Verification CA Test Implementation"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(drcHelper)
source("../knitr-setup.R")
```

## Comparison: Cochran-Armitage Test Implementation vs `prop.trend.test`

The implementation in **drcHelper** and `prop.trend.test` are mathematically equivalent but use different computational approaches. In particular, the drcHelper implementation uses actual dose values as default scoring, adds overdispersion handling by Rao-Scott correction for clustered data, the grouping by doses are done automatically, and uses the more intuitive Z-statistic. It is more comprehensive and has a cleaner interface for dose-response studies.

Here's the detailed comparison:


## Key Differences 

| Aspect | drcHelper Implementation | `prop.trend.test` |
|--------|-------------------|-------------------|
| **Computational Method** | Direct Z-statistic calculation | Weighted linear regression |
| **Test Statistic** | Z-score (normal distribution) | Chi-squared ($Z^2$) |
| **Score Usage** | Uses actual `doses` values | Uses `seq_along(x)` by default |
| **Overdispersion** | Rao-Scott correction available | No correction |
| **Data Grouping** | Groups by doses automatically | Assumes pre-grouped data |

## Mathematical Relationship

The relationship between the two approaches:

- **Z-statistic**: $Z = \frac{\sum n_i(s_i - \bar{s})p_i}{\sqrt{p(1-p)\sum n_i(s_i - \bar{s})^2}}$
- **prop.trend.test Chi-squared**: $\chi^2 = Z^2$

Where $p(\chi^2) = 2 \times p(|Z|)$ for two-sided tests.



## Verification Example ✅

```{r}
# Test data
successes <- c(83, 90, 129, 70)
totals <- c(86, 93, 136, 82)

# R's function (default scores: 1,2,3,4)
result_r <- prop.trend.test(successes, totals)
result_r
# drcHelper function (after correction, using doses = 1,2,3,4)
result <- cochranArmitageTrendTest(successes, totals, doses = 1:4)
result
# Should give: sqrt(result_r$statistic) ≈ abs(result_yours$statistic)
sqrt(result_r$statistic) - abs(result$statistic)
```





