---
title: "Limitations of Williams' Test"
format: html
---


We first creates simulated data with exaggerated non-monotonicity and variability to clearly illustrate the problem:

```{r}
# ── Simulated data illustrating Williams' test limitations ──
# Mimics OECD 219 development rate: 6 treatments, 4 replicates
# Deliberately non-monotonic: hormetic bump at low doses, dip at mid-dose

set.seed(2026)
trt_levels <- c("Control", "6.25", "12.5", "25", "50", "100")

# True pattern: hormesis (stimulation at low dose) + effect at high dose
true_means <- c(
  Control = 0.0520,
  `6.25`  = 0.0540,   # ← stimulation (higher than control)
  `12.5`  = 0.0545,   # ← stimulation continues
  `25`    = 0.0530,   # ← starts declining
  `50`    = 0.0480,   # ← clear effect
  `100`   = 0.0500    # ← partial recovery (non-monotonic)
)

# Generate replicate data with realistic within-group SD (~0.002)
sd_within <- 0.002

df_fake <- purrr::imap_dfr(true_means, \(mu, dose) {
  tibble::tibble(
    Dose      = factor(dose, levels = trt_levels),
    Replicate = 1:4,
    Response  = rnorm(4, mean = mu, sd = sd_within)
  )
})

# ── 1. Visualise the raw data ──
library(ggplot2)

group_summary <- df_fake |>
  dplyr::group_by(Dose) |>
  dplyr::summarise(
    mean = mean(Response),
    se   = sd(Response) / sqrt(dplyr::n()),
    n    = dplyr::n(),
    .groups = "drop"
  )

p_raw <- ggplot(group_summary, aes(x = Dose, y = mean)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean - 1.96 * se, ymax = mean + 1.96 * se),
                width = 0.2) +
  geom_point(data = df_fake, aes(y = Response),
             alpha = 0.4, size = 2, color = "steelblue") +
  labs(
    title = "Simulated Development Rate — observed means",
    subtitle = "Non-monotonic pattern: hormesis + partial recovery",
    y = "Development rate (1/day)"
  ) +
  theme_bw()

# ── 2. Run PAVA and show what Williams sees ──
pava_decreasing <- function(means, weights) {
  n <- length(means)
  block_val <- means
  block_wt  <- weights
  block_id  <- as.list(seq_len(n))
  
  log <- list()
  iter <- 0
  
  repeat {
    violation <- NA
    for (i in seq_len(length(block_val) - 1)) {
      if (block_val[i] < block_val[i + 1]) {
        violation <- i
        break
      }
    }
    if (is.na(violation)) break
    
    iter <- iter + 1
    i <- violation
    new_val <- (block_val[i] * block_wt[i] + block_val[i + 1] * block_wt[i + 1]) /
      (block_wt[i] + block_wt[i + 1])
    
    log[[iter]] <- list(
      iteration = iter,
      merged    = c(block_id[[i]], block_id[[i + 1]]),
      from      = c(block_val[i], block_val[i + 1]),
      to        = new_val
    )
    
    block_val  <- c(block_val[-(i:(i+1))],  NA)  # placeholder
    block_wt   <- c(block_wt[-(i:(i+1))],   NA)
    block_id   <- c(block_id[-(i:(i+1))],   list(NULL))
    # Insert merged block at position i
    block_val  <- append(block_val[-length(block_val)], new_val, after = i - 1)
    block_wt   <- append(block_wt[-length(block_wt)], block_wt[i] + block_wt[i + 1], after = i - 1)
    # Fix: recompute properly
    block_val  <- block_val  # already correct
    block_wt[i] <- sum(unlist(log[[iter]]$merged) |> length()) * weights[1]  
    block_id   <- append(block_id[-length(block_id)], 
                         list(unlist(log[[iter]]$merged)), after = i - 1)
  }
  
  # Expand back to original group positions
  iso_means <- rep(NA_real_, length(means))
  for (j in seq_along(block_val)) {
    iso_means[unlist(block_id[[j]])] <- block_val[j]
  }
  
  list(isotonised = iso_means, n_iterations = iter, log = log)
}

# ── Reusable function: PAVA summary table + plot ──
summarise_pava <- function(group_summary, title_suffix = "") {
  obs_means <- group_summary$mean
  iso_fit   <- isoreg(-obs_means)
  iso_means <- -iso_fit$yf

  # Comparison table
  comparison <- group_summary |>
    dplyr::mutate(
      observed_mean   = mean,
      isotonised_mean = iso_means,
      changed         = abs(observed_mean - isotonised_mean) > 1e-10,
      pct_change      = (isotonised_mean - observed_mean) / observed_mean * 100
    ) |>
    dplyr::select(Dose, n, observed_mean, isotonised_mean, changed, pct_change)

  # Identify PAVA blocks (runs of identical isotonised means)
  n_changed  <- sum(comparison$changed)
  n_blocks   <- length(rle(iso_means)$values)

  # Plot data
  plot_data <- group_summary |>
    dplyr::mutate(isotonised = iso_means) |>
    tidyr::pivot_longer(cols = c(mean, isotonised),
                        names_to = "type", values_to = "value") |>
    dplyr::mutate(
      type = dplyr::if_else(type == "mean", "Observed", "PAVA (isotonised)")
    )

  p <- ggplot(plot_data, aes(x = Dose, y = value, color = type, group = type)) +
    geom_point(size = 3) +
    geom_line() +
    scale_color_manual(
      values = c("Observed" = "black", "PAVA (isotonised)" = "red")
    ) +
    labs(
      title    = paste0("Observed vs PAVA-isotonised means", title_suffix),
      subtitle = sprintf(
        "%d of %d means changed; %d blocks after amalgamation",
        n_changed, nrow(comparison), n_blocks
      ),
      y     = "Development rate (1/day)",
      color = NULL
    ) +
    theme_bw() +
    theme(legend.position = "top")

  list(comparison = comparison, plot = p,
       n_changed = n_changed, n_blocks = n_blocks)
}

# ── Apply to Scenario A ──
pava_a <- summarise_pava(group_summary, title_suffix = " — Scenario A")

p_raw
pava_a$plot
pava_a$comparison
```



The simulated data has two features that break monotonicity:

1. **Hormesis** at 6.25 and 12.5 mg/L — response *increases* above control before declining
2. **Partial recovery** at 100 mg/L — the highest dose has a higher response than 50 mg/L

PAVA will amalgamate multiple groups to force monotonicity.


## Williams' step-down procedure

Williams' test uses a **step-down procedure starting from the highest dose**:

1. Compare the highest dose (100) vs Control using the PAVA-isotonised mean
2. If significant → reject, move to the next lower dose (50)
3. If significant → reject, move to 25
4. Continue until a dose is **not** significant → **stop**. That dose is the NOEC, and all lower doses are automatically accepted (not tested)



- **50 mg/L** has a genuine effect (low observed mean), but PAVA pools it with 100 mg/L (which recovered), pulling the isotonised mean *up*
- This can make the 50 mg/L comparison *less* significant than it should be, because the amalgamated mean is diluted by the recovery at 100
- Conversely, the **100 mg/L** comparison uses that same pooled mean, so it may appear more affected than the raw data suggests



| Limitation | What happens |
|-----------|-------------|
| **Loss of group identity** | Multiple dose groups get the same isotonised mean |
| **Biased estimates** | Amalgamated means can be far from any individual group's true mean |
| **Hormesis masked** | Stimulatory effects at low doses are averaged away with Control |
| **Diluted effects** | A genuine effect at dose $k$ can be masked if PAVA pools it with a recovered dose $k+1$ |
| **Step-down stops early** | If the highest dose is non-significant (e.g., due to recovery), the step-down stops immediately → all doses are declared non-significant, even if mid-range doses have real effects |
| **Reduced specificity** | Cannot determine which specific dose is the LOEC within an amalgamated block |

The **step-down stopping** point is the most critical issue for this dataset: if the PAVA-pooled {50+100} mean is not significant against Control, Williams would declare NOEC ≥ 100 and never even test 50 mg/L individually — despite 50 mg/L having a statistically significant effect under Dunnett (p = 0.025).


## Fragility of the monotonicity pre-test

ToxRat's decision tree uses `drcHelper::monotonicityTest()` — a linear + quadratic contrast on ranks — to decide whether Williams or Dunnett is appropriate. The logic is:

- **Linear significant + Quadratic not significant** → monotonic → Williams
- **Quadratic significant** → non-monotonic → Dunnett

But this pre-test has limited power with small sample sizes (typically $n = 4$ replicates). We can demonstrate this fragility by applying multiple trend tests to the same simulated data.

```{r}
#| label: monotonicity-pretest

# drcHelper monotonicity test (linear + quadratic contrasts on ranks)
mono_result <- drcHelper::monotonicityTest(
  Data      = as.data.frame(df_fake),
  Treatment = "Dose",
  Response  = "Response"
)

mono_result
```

The linear contrast may detect the net downward slope, but the quadratic contrast — which should flag the non-monotonic bump — often lacks power to reach significance.

### Corroborating trend tests

We apply three additional tests to assess whether the evidence for a monotone trend is robust:

```{r}
#| label: trend-tests

library(tukeytrend)
library(multcomp)

# Numeric dose for trend models
df_trend <- as.data.frame(df_fake) |>
  dplyr::mutate(
    DoseNum = dplyr::if_else(
      as.character(Dose) == "Control", 0,
      suppressWarnings(as.numeric(as.character(Dose)))
    )
  )

# 1. Tukey Trend Test (multiple dose scalings, multiplicity-adjusted)
fit_lm <- lm(Response ~ DoseNum, data = df_trend)

tt_fit <- tukeytrendfit(
  fit_lm,
  dose    = "DoseNum",
  scaling = c("ari", "ord", "arilog"),
  ddf     = "residual"
)

tt_result <- summary(glht(
  model       = tt_fit$mmm,
  linfct      = tt_fit$mlf,
  alternative = "less"
))

# 2. Jonckheere-Terpstra (nonparametric ordered alternative)
jt_result <- PMCMRplus::jonckheereTest(
  Response ~ Dose,
  data        = df_trend,
  alternative = "less"
)

# 3. Compile comparison
tt_pvals <- tt_result$test$pvalues
tt_names <- names(tt_result$test$tstat)

trend_comparison <- tibble::tibble(
  Test = c(
    "monotonicityTest — Linear",
    "monotonicityTest — Quadratic",
    paste0("Tukey Trend — ", c("ari", "ord", "arilog")),
    "Jonckheere-Terpstra"
  ),
  p_value = round(c(
    mono_result$`Pr(>|t|)`[1],
    mono_result$`Pr(>|t|)`[2],
    tt_pvals,
    jt_result$p.value
  ), 4),
  Significant = dplyr::if_else(p_value < 0.05, "Yes", "No")
)

trend_comparison
```

This table typically reveals inconsistency: the linear contrast may be significant while the Tukey ordinal scaling is not, or vice versa. When trend tests disagree about whether a monotone trend exists, the decision between Williams and Dunnett becomes **arbitrary** — it depends on which test you chose as gatekeeper. This is a fundamental fragility in the two-stage procedure.

### Scenario A summary

In the data above (seed 2026), the monotonicity pre-test finds the linear contrast **borderline non-significant** (p ≈ 0.052) — ToxRat would select Dunnett. Yet the Tukey trend test detects a significant decreasing trend under all three dose scalings (p < 0.025). The tests contradict each other.

### Scenario B: Linear significant, Tukey inconsistent

We now generate a second dataset where the *opposite* happens: `monotonicityTest` finds a significant linear trend (borderline, p ≈ 0.03–0.04), but the Tukey trend test is split — only some dose scalings reach significance.

```{r}
#| label: scenario-b-data

# Scenario B: very mild monotonic decline with a bump at one mid-dose
# The overall slope is weak enough that the linear contrast is borderline,
# while the bump at 12.5 disrupts ordinal/arilog scalings.

set.seed(42)

true_means_b <- c(
  Control = 0.0525,
  `6.25`  = 0.0523,
  `12.5`  = 0.0535,   # ← bump (higher than both neighbours)
  `25`    = 0.0520,
  `50`    = 0.0518,
  `100`   = 0.0515
)

df_fake_b <- purrr::imap_dfr(true_means_b, \(mu, dose) {
  tibble::tibble(
    Dose      = factor(dose, levels = trt_levels),
    Replicate = 1:4,
    Response  = rnorm(4, mean = mu, sd = 0.0018)
  )
})

# Visualise
group_summary_b <- df_fake_b |>
  dplyr::group_by(Dose) |>
  dplyr::summarise(
    mean = mean(Response),
    se   = sd(Response) / sqrt(dplyr::n()),
    n    = dplyr::n(),
    .groups = "drop"
  )

ggplot(group_summary_b, aes(x = Dose, y = mean)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean - 1.96 * se, ymax = mean + 1.96 * se),
                width = 0.2) +
  geom_point(data = df_fake_b, aes(y = Response),
             alpha = 0.4, size = 2, color = "steelblue") +
  labs(
    title = "Scenario B — very mild decline with mid-dose bump",
    subtitle = "Weak overall trend; bump disrupts ordinal/arilog scalings",
    y = "Development rate (1/day)"
  ) +
  theme_bw()

# PAVA summary for Scenario B
pava_b <- summarise_pava(group_summary_b, title_suffix = " — Scenario B")
pava_b$plot
pava_b$comparison
```

```{r}
#| label: scenario-b-trend-tests

# Monotonicity pre-test
mono_b <- drcHelper::monotonicityTest(
  Data      = as.data.frame(df_fake_b),
  Treatment = "Dose",
  Response  = "Response"
)

# Tukey trend test
df_trend_b <- as.data.frame(df_fake_b) |>
  dplyr::mutate(
    DoseNum = dplyr::if_else(
      as.character(Dose) == "Control", 0,
      suppressWarnings(as.numeric(as.character(Dose)))
    )
  )

fit_lm_b <- lm(Response ~ DoseNum, data = df_trend_b)

tt_fit_b <- tukeytrendfit(
  fit_lm_b,
  dose    = "DoseNum",
  scaling = c("ari", "ord", "arilog"),
  ddf     = "residual"
)

tt_result_b <- summary(glht(
  model       = tt_fit_b$mmm,
  linfct      = tt_fit_b$mlf,
  alternative = "less"
))

# Jonckheere-Terpstra
jt_b <- PMCMRplus::jonckheereTest(
  Response ~ Dose,
  data        = df_trend_b,
  alternative = "less"
)

# Comparison table
tt_pvals_b <- tt_result_b$test$pvalues
tt_names_b <- names(tt_result_b$test$tstat)

trend_comparison_b <- tibble::tibble(
  Test = c(
    "monotonicityTest — Linear",
    "monotonicityTest — Quadratic",
    paste0("Tukey Trend — ", c("ari", "ord", "arilog")),
    "Jonckheere-Terpstra"
  ),
  p_value = round(c(
    mono_b$`Pr(>|t|)`[1],
    mono_b$`Pr(>|t|)`[2],
    tt_pvals_b,
    jt_b$p.value
  ), 4),
  Significant = dplyr::if_else(p_value < 0.05, "Yes", "No")
)

trend_comparison_b
```

### Two scenarios, opposite conclusions

| | Scenario A (seed 2026) | Scenario B (seed 42) |
|--|----------------------|---------------------|
| **Data pattern** | Hormesis + recovery | Very mild decline + mid-dose bump |
| **monotonicityTest Linear** | Not significant (p ≈ 0.052) | Borderline / depends on realisation |
| **Tukey Trend** | All significant | Weakened by bump (depends on scaling) |
| **ToxRat decision** | → Dunnett | Ambiguous |
| **Reality** | Data is clearly non-monotonic | Data is also non-monotonic |

The same underlying problem (non-monotonicity) leads to opposite gatekeeper decisions depending on the specific data pattern and random variation. This demonstrates that the two-stage approach (pre-test monotonicity → choose test) is fragile and can be misleading.


## Comparison: Williams vs Dunnett (one-sided and two-sided)

Now we run all three tests on both simulated datasets and compare their conclusions.

### Scenario A

```{r}
#| label: williams-vs-dunnett-a

# Williams' test (uses PAVA-isotonised means, step-down from highest dose)
williams_res <- drcHelper::broom_williams(
  Response ~ Dose,
  data      = as.data.frame(df_fake),
  method    = "Williams_JG",
  direction = "decreasing"
)

# Dunnett's test — one-sided (decreasing)
dunnett_1s <- drcHelper::dunnett_test(
  data                  = as.data.frame(df_fake),
  response_var          = "Response",
  dose_var              = "Dose",
  tank_var              = "Replicate",
  include_random_effect = FALSE,
  alternative           = "less"
)

# Dunnett's test — two-sided
dunnett_2s <- drcHelper::dunnett_test(
  data                  = as.data.frame(df_fake),
  response_var          = "Response",
  dose_var              = "Dose",
  tank_var              = "Replicate",
  include_random_effect = FALSE,
  alternative           = "two.sided"
)

# ── Build clean comparison table ──
dose_levels <- c("6.25", "12.5", "25", "50", "100")

tbl_williams <- williams_res |>
  dplyr::transmute(
    Dose      = stringr::str_extract(comparison, "^[\\d.]+"),
    Test      = "Williams",
    Estimate  = round(estimate, 6),
    Statistic = round(`t'-stat`, 3),
    p_value   = NA_character_,
    Decision  = decision
  )

tbl_dunnett_1s <- dunnett_1s$results_table |>
  dplyr::transmute(
    Dose      = stringr::str_extract(comparison, "^[\\d.]+"),
    Test      = "Dunnett (one-sided)",
    Estimate  = round(estimate, 6),
    Statistic = round(statistic, 3),
    p_value   = format(round(p.value, 4), nsmall = 4),
    Decision  = dplyr::if_else(significant, "reject", "accept")
  )

tbl_dunnett_2s <- dunnett_2s$results_table |>
  dplyr::transmute(
    Dose      = stringr::str_extract(comparison, "^[\\d.]+"),
    Test      = "Dunnett (two-sided)",
    Estimate  = round(estimate, 6),
    Statistic = round(statistic, 3),
    p_value   = format(round(p.value, 4), nsmall = 4),
    Decision  = dplyr::if_else(significant, "reject", "accept")
  )

comparison_table <- dplyr::bind_rows(
  tbl_williams, tbl_dunnett_1s, tbl_dunnett_2s
) |>
  dplyr::arrange(
    factor(Dose, levels = dose_levels),
    factor(Test, levels = c("Williams", "Dunnett (one-sided)", "Dunnett (two-sided)"))
  )

comparison_table
```

```{r}
#| label: noec-helper

# Helper: derive NOEC from Williams step-down result
derive_noec_williams <- function(williams_res) {
  if (is.null(williams_res) || nrow(williams_res) == 0) return(NA)
  decisions <- williams_res$decision
  doses_chr <- gsub(" - 0.*", "", williams_res$comparison)
  if (all(decisions == "accept")) return(paste(">=", doses_chr[length(doses_chr)]))
  if (all(decisions == "reject")) return(paste("<", doses_chr[1]))
  reject_idx <- which(decisions == "reject")
  first_reject <- min(reject_idx)
  if (first_reject == 1) return(paste("<", doses_chr[1]))
  doses_chr[first_reject - 1]
}
```

```{r}
#| label: noec-summary-a

# Extract NOEC from each test
noec_williams <- derive_noec_williams(williams_res)
noec_dunnett_1s <- dunnett_1s$noec
noec_dunnett_2s <- dunnett_2s$noec

tibble::tibble(
  Test = c("Williams", "Dunnett (one-sided)", "Dunnett (two-sided)"),
  NOEC = c(noec_williams, noec_dunnett_1s, noec_dunnett_2s)
)
```

### Scenario B

```{r}
#| label: williams-vs-dunnett-b

williams_res_b <- drcHelper::broom_williams(
  Response ~ Dose,
  data      = as.data.frame(df_fake_b),
  method    = "Williams_JG",
  direction = "decreasing"
)

dunnett_1s_b <- drcHelper::dunnett_test(
  data                  = as.data.frame(df_fake_b),
  response_var          = "Response",
  dose_var              = "Dose",
  tank_var              = "Replicate",
  include_random_effect = FALSE,
  alternative           = "less"
)

dunnett_2s_b <- drcHelper::dunnett_test(
  data                  = as.data.frame(df_fake_b),
  response_var          = "Response",
  dose_var              = "Dose",
  tank_var              = "Replicate",
  include_random_effect = FALSE,
  alternative           = "two.sided"
)

tbl_williams_b <- williams_res_b |>
  dplyr::transmute(
    Dose      = stringr::str_extract(comparison, "^[\\d.]+"),
    Test      = "Williams",
    Estimate  = round(estimate, 6),
    Statistic = round(`t'-stat`, 3),
    p_value   = NA_character_,
    Decision  = decision
  )

tbl_dunnett_1s_b <- dunnett_1s_b$results_table |>
  dplyr::transmute(
    Dose      = stringr::str_extract(comparison, "^[\\d.]+"),
    Test      = "Dunnett (one-sided)",
    Estimate  = round(estimate, 6),
    Statistic = round(statistic, 3),
    p_value   = format(round(p.value, 4), nsmall = 4),
    Decision  = dplyr::if_else(significant, "reject", "accept")
  )

tbl_dunnett_2s_b <- dunnett_2s_b$results_table |>
  dplyr::transmute(
    Dose      = stringr::str_extract(comparison, "^[\\d.]+"),
    Test      = "Dunnett (two-sided)",
    Estimate  = round(estimate, 6),
    Statistic = round(statistic, 3),
    p_value   = format(round(p.value, 4), nsmall = 4),
    Decision  = dplyr::if_else(significant, "reject", "accept")
  )

comparison_table_b <- dplyr::bind_rows(
  tbl_williams_b, tbl_dunnett_1s_b, tbl_dunnett_2s_b
) |>
  dplyr::arrange(
    factor(Dose, levels = dose_levels),
    factor(Test, levels = c("Williams", "Dunnett (one-sided)", "Dunnett (two-sided)"))
  )

comparison_table_b
```

```{r}
#| label: noec-summary-b

noec_williams_b <- derive_noec_williams(williams_res_b)
noec_dunnett_1s_b <- dunnett_1s_b$noec
noec_dunnett_2s_b <- dunnett_2s_b$noec

tibble::tibble(
  Test = c("Williams", "Dunnett (one-sided)", "Dunnett (two-sided)"),
  NOEC = c(noec_williams_b, noec_dunnett_1s_b, noec_dunnett_2s_b)
)
```

### Key observations

1. **Williams uses amalgamated estimates**: doses within the same PAVA block share the same estimate, so their test statistics are identical — the test cannot distinguish among them.

2. **Dunnett one-sided vs two-sided**: the one-sided test has more power (lower p-values) because it tests only $H_1: \mu_i < \mu_0$. The two-sided test is more conservative but can also detect stimulatory effects at low doses.

3. **Step-down vs per-dose**: Williams' step-down can miss mid-range effects if the highest dose has recovered (PAVA dilutes the effect). Dunnett tests each dose independently.

4. **NOEC discrepancies**: when the dose-response is non-monotonic, the three tests may yield different NOECs — the choice of test drives the regulatory conclusion.

5. **Scenario comparison**: Scenario A (hormesis + partial recovery) and Scenario B (mild decline + mid-dose bump) illustrate that small changes in the shape of the dose–response can flip which tests are significant and which NOECs result. Monotonicity pre-tests and trend tests may disagree, and the Williams vs Dunnett NOEC can diverge — all depending on the specific non-monotonic pattern present in the data.


## Notes

### Terminology

**Amalgamated mean** = the pooled weighted average produced by PAVA. When two adjacent groups violate the monotonicity constraint, their means are "amalgamated" (merged) into a single value. The term comes from the original Williams (1971, 1972) papers. 

**Isotonised** = forced to satisfy an isotonic (monotone) constraint. "Isotonic regression" is the general statistical term for fitting a monotone function to data by minimizing squared error subject to an ordering constraint. PAVA is the algorithm that solves this. The isotonised means are the PAVA output — the closest monotone sequence to the observed means.

So: **amalgamated mean = isotonised mean = PAVA-adjusted mean**. 