---
title: "KNF Algae Growth Inhibition "
author: "Zhenglei Gao"
editor_options: 
  chunk_output_type: console
---


```{r}
library(tidyverse)
n<- c(12,rep(6,6))
a<-qt(1-0.05/2, n-1)
a
a <- qnorm(1-0.05/2)
a
dat <- structure(list(Treatment = structure(c(1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 2L, 2L, 2L, 2L, 3L, 3L, 3L, 3L, 4L, 4L, 
4L, 4L, 5L, 5L, 5L, 5L, 6L, 6L, 6L, 6L, 7L, 7L, 7L, 7L), levels = c("Control", 
"0.0015", "0.004", "0.016", "0.05", "0.16", "0.5"), class = c("ordered", 
"factor")), Response = c(0.0500751550402492, 0.0549766102708876, 
0.0510179538436234, 0.0504660348080099, 0.0499809345685877, 0.0539911129987799, 
0.054056210976094, 0.0546848380887881, 0.0511028700289316, 0.0521132294706069, 
0.0516136666906998, 0.054006668312382, 0.0584551412709057, 0.0537452668039128, 
0.0523058493653312, 0.0597996499408036, 0.0551921754687093, 0.0549841190670617, 
0.0557558414385654, 0.0571090072982945, 0.0550790786321275, 0.0538856068123132, 
0.0566810136275366, 0.0530885395532474, 0.0538114115982316, 0.0511170609363355, 
0.0513881320494227, 0.0505882840408944, 0.0232783194745891, 0.0119942272946239, 
0.0275260484650731, 0.0305045980902389, 0.00691141055684537, 
0.000590934835374355, 0.0072092273690179, 0.00127559194218367
)), row.names = c(NA, -36L), class = "data.frame")

## write.csv(dat,file="JT_dat1.csv",row.names = F) 
# Create the dataset
datsum <-dat %>% group_by(Treatment) %>% summarise(mean=mean(Response), SE=sd(Response)/sqrt(length(Response))) %>% mutate(Lower=mean-a*SE,Upper=mean+a*SE)
```



```{r}

theme_set(theme_bw())
ggplot(dat,aes(x=Treatment,y=Response)) + geom_point() + geom_point(data = datsum,aes(as.numeric(Treatment)+0.2,y=mean),position=position_dodge(width = 0.9),col = "red",size = 3,shape = 24, fill = "pink")+
geom_errorbar(data = datsum, aes(x=as.numeric(Treatment)+0.2,y=mean,ymin = Lower, ymax = Upper),col="red",position=position_dodge(width = 0.9),width = .2)

ggsave("JT_example.png",width = 5 ,height =3.5,dpi=300 )
ggsave("JT_example.pdf",width = 5 ,height =3.5,dpi=300 )
```

- https://www.tandfonline.com/doi/full/10.1080/03610920601041556#:~:text=This%20article%20examines%20the%20variance


Jonckheere's test and Kendall's tau) are closely related. In some software implementations, the two test produce identical p-values.


```{r}
## Jonckheere, A. R. (1954). A distribution-free k-sample test again ordered alternatives. Biometrika 41:133-145.

## Terpstra, T. J. (1952). The asymptotic normality and consistency of Kendall's test against trend, when ties are present in one ranking. Indagationes Mathematicae 14:327-333.

DescTools::JonckheereTerpstraTest(Response~Treatment,data=dat)
DescTools::JonckheereTerpstraTest(Response~Treatment,data=dat%>% filter(Treatment!="0.5"))
DescTools::JonckheereTerpstraTest(Response~Treatment,data=dat%>% filter(Treatment!="0.5" & Treatment!="0.16"))
DescTools::JonckheereTerpstraTest(Response~Treatment,data=dat%>% filter(Treatment!="0.5" & Treatment!="0.16" & Treatment!="0.05"))
DescTools::JonckheereTerpstraTest(Response~Treatment,data=dat%>% filter(Treatment!="0.5" & Treatment!="0.16" & Treatment!="0.05" & Treatment != "0.016"))

DescTools::JonckheereTerpstraTest(Response~Treatment,data=dat%>% filter(Treatment!="0.5" & Treatment!="0.16" & Treatment!="0.05" & Treatment != "0.016" & Treatment != "0.004" ))


PMCMRplus::jonckheereTest(dat$Response,dat$Treatment)
cor.test(dat$Response,as.numeric(dat$Treatment),method="kendall",exact = T)

clinfun::jonckheere.test(dat$Response,dat$Treatment)
## example from Exact Test (Mehta), figure 12.2
dat1 <- matrix(
  c(100, 18, 50, 50, 1, 1, 1, 1, 0, 1, 1, 1, 0, 0, 1, 1), 4,
  dimnames = list(dose = 1:4 * 100,
                  toxicity = c('Mild', 'Moderate', 'Severe', 'Death'))
)
rawr::jt.test(dat1)

tbl <- table(mtcars$gear, mtcars$cyl)
rawr::jt.test(tbl)
## compare
cor.test(mtcars$gear, mtcars$cyl, method = 'kendall', exact = FALSE)
clinfun::jonckheere.test(mtcars$gear, mtcars$cyl)
```

```{r}
PMCMRplus::stepDownTrendTest(Response~Treatment,data=dat,test = "jonckheere")
library(PMCMRplus)
welchManyOneTTest(Response~Treatment,data=dat,p.adjust.method = "none")
welchManyOneTTest(Response~Treatment,data=dat,p.adjust.method = "holm",alternative = "less")
welchManyOneTTest(Response~Treatment,data=dat,p.adjust.method = "holm")
```


```{r}
library(emmeans)
mod0 <- lm(Response~Treatment,data=dat)
treat.means <- emmeans(mod0, ~Treatment)
treat.means

contrast(treat.means, adjust="none", method="dunnett", ref = 1)
contrast(treat.means, method="dunnett", ref = 1)

DescTools::DunnettTest(Response~Treatment,data=dat)
library(multcomp)
summary(glht(mod0,mcp(Treatment="Dunnett"))) %>% broom::tidy() %>% dplyr::select(-c(term,null.value)) %>% knitr::kable(digits = 3) %>% kableExtra::kable_classic() ## Same as DescTools implementation
```


glht(..., mcp(trt = "Dunnett")) pulls its P-values out of a multivariate t distribution. They call mvtnorm::pmvt with the observed t statistics and correlation matrix (actual code here). DescTools::DunnettTest does roughly the same (code here).

emmeans does something.. different. Already in the documentation they call their dunnetx method "a close approximation to the Dunnett adjustment". This is the actual source:



```{r eval=F}
### My approximate dunnett distribution 
### - a mix of the Tukey cdf and Sidak-corrected t
.pdunnx = function(x, k, df, twt = (k - 1)/k) {
    tukey = ptukey(sqrt(2)*x, (1 + sqrt(1 + 8*k))/2, df)
    sidak = (pf(x^2, 1, df))^k
    twt*tukey + (1 - twt)*sidak
}
```

(x are the t statistics, k the number of contrasts and df self-explanatory)

You can make emmeans match the others by using adjust="mvt" which will then also call mvtnorm::pmvt. Either way I wouldn't sweat the small differences in P-values, or I'd worry equally much about the fact that any default call to mvtnorm::pmvt (which wasn't changed in any of the above) targets an absolute epsilon of 1e−3
, so your P-values would only be accurate to about 3 decimal places anyway.

Small aside, unless you include a seed there will never be a match against any subsequent runs of your randomly sampling code (making your example non-reproducible).

```{r}
library(nlme)

gls0 <- gls(Response ~ Treatment, data=dat,weights=varIdent(form= ~1|Treatment))
plot(gls0,Response ~ fitted(.) | Treatment)
plot(gls0,Treatment ~ fitted(.))
plot(gls0,Treatment ~ resid(.),abline=0)
dat$fitted <- fitted(gls0)
ggplot(dat,aes(x=fitted,y=Response,col=Treatment))+geom_point()+geom_abline(intercept = 0,slope=1)
plot(gls0, resid(.) ~ fitted(.) | Treatment, abline = 0)
treat.means <- emmeans(gls0, ~ Treatment)
contrast(treat.means, adjust="none", method="dunnett", ref = 1)

summary(glht(gls0,mcp(Treatment="Dunnett"),alternative = "less")) 
```

```{r}
testres1 <- summary(glht(gls0,mcp(Treatment="Dunnett"))) 
testres1 %>% broom::tidy() %>% dplyr::select(-c(term,null.value)) %>% knitr::kable(digits = 3) %>% kableExtra::kable_classic()
```

```{r}
testres2 <- summary(glht(gls0,mcp(Treatment="Dunnett"),alternative="less")) 
testres2 %>% broom::tidy() %>% dplyr::select(-c(term,null.value)) %>% knitr::kable(digits = 3) %>% kableExtra::kable_classic()
```

## Manipulate the dataset

Increase the variance at NOEC 0.05 by changing the maximum or the minimum.

```{r}
library(multcomp)
mod0 <- lm(Response~Treatment,data=dat)
summary(glht(mod0,linfct = mcp(Treatment="Dunnett"),alternative="less"))
dat2 <- dat
dat2$Response[dat$Treatment==0.05] <- c(0.0538114115982316, 0.0511170609363355, 0.0513881320494227, 
0.02)
mod2 <- lm(Response~Treatment,data=dat2)
summary(glht(mod2,linfct = mcp(Treatment="Dunnett"),alternative="less"))
PMCMRplus::stepDownTrendTest(Response~Treatment,data=dat2,test="jonckheere")
dat3 <- dat
dat3$Response[dat$Treatment==0.05] <-c(0.1, 0.0511170609363355, 0.0513881320494227, 
0.0505882840408944)
mod3 <- lm(Response~Treatment,data=dat3)
summary(glht(mod3,linfct = mcp(Treatment="Dunnett"),alternative="less"))
PMCMRplus::stepDownTrendTest(Response~Treatment,data=dat3,test="jonckheere")

```

